---
title: "deposits"
output:
  md_document:
    variant: gfm

  rmarkdown::html_vignette:
    self_contained: no
---

# deposits

<!-- badges: start -->
[![R build status](https://github.com/ropenscilabs/deposits/workflows/R-CMD-check/badge.svg)](https://github.com/ropenscilabs/deposits/actions?query=workflow%3AR-CMD-check)
[![codecov](https://codecov.io/gh/ropenscilabs/deposits/branch/main/graph/badge.svg)](https://codecov.io/gh/ropenscilabs/deposits)
[![Project Status: Concept](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
<!-- badges: end -->


The `deposits` R package is a universal client for depositing and accessing
research data anywhere. Currently supported services are
[zenodo](https://zenodo.org) and [figshare](https://figshare.com). Instructions
for installing and setting up the package are in [a separate
vignette](https://docs.ropensci.org/deposits/articles/install-setup.html). This
README gives a brief overview of package usage, with more detail in [the
introductory
vignette](https://docs.ropensci.org/deposits/articles/deposits.html).


## Data Repositories

The list of data repositories currently supported is accessible by the
`deposits_services()` function:

```{r services}
library (deposits)
deposits_services ()
```

[`zenodo`](https://zenodo.org) offers a "sandbox" environment, which
offers an ideal environment for testing the functionality of this
package.


## The deposits workflow

The `deposits` package uses [the `R6` package](https://github.com/r-lib/R6) to
create [a
`depositsClient`](https://docs.ropensci.org/deposits/reference/depositsClient.html)
used to perform all functions. All `deposits` operations start with a client
constructed with the `new()` function:

```{r client-new, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
cli
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>    hostdata : <none>
#>    metadata : <none>
```

Having constructed a new client, the deposits workflow then involves the
following main steps:

1. Fill the client with metadata;
2. Create a new deposit on the nominated service; and
3. Upload files for the deposit.

### Metadata

Metadata are data describing a deposit, while "data" are the deposit itself,
generally uploaded as files as described below. Metadata include details on
authors, titles, file types, descriptions, and many other terms. These can be
included in a deposits client in several ways, and either on initial
construction of an client, or as a separate step applied to an existing client.

One of the easiest ways to specify metadata is as a list of named terms like
the following example:

```{r metadata}
metadata <- list (
    title = "New Title",
    abstract = "This is the abstract",
    creator = list ("A. Person", "B. Person")
)
```

These data can be used in construction of a new client by passing a `metadata`
argument:

```{r client-new-metadata, eval = FALSE}
cli <- depositsClient$new (
    service = "zenodo",
    sandbox = TRUE,
    metadata = metadata
)
cli
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>    hostdata : <none>
#>    metadata : 3 terms (see 'metadata' element for details)
```

Equivalently, they can be added to an existing client with [the
`deposit_fill_metadata()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_fill_metadata):

```{r client-fill-metadata, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
cli$deposit_fill_metadata (metadata)
```

Note that R6 functions are called directly on the client, with the object
itself (`cli`) updated by the call. Other ways of specifying and entering
metadata are described in [the introductory
vignette](https://docs.ropensci.org/deposits/articles/deposits.html).

### Create a new deposit

The metadata filled with the above steps can then be used to initiate a new
deposit on the associated server using the `deposit_new()` function:

```{r deposit_new, eval = FALSE}
cli$deposit_new ()
cli
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>  url_service : https://sandbox.zenodo.org/deposit/1065666
#>   deposit id : 1065666
#>     hostdata : list with 14  elements
#>     metadata : 11 terms (see 'metadata' element for details)
```

The client now includes several additional elements, notably a "deposit id"
(stored in `cli$id`) giving the unique identifier for the new deposit, and a
`hostdata` item with, in this case,  14 elements as specified by the host
service. The `url_service` is the URL for the newly-created deposit. (Viewing
in a web browser will require logging in for all private and sandbox deposits).

### Uploading (and downloading) files

A deposit is really about data, not just metadata. Uploading data to a deposit
is as simple as the following function:

```{r deposit_upload, eval = FALSE}
cli$deposit_upload_file (path = "<path>/<to>/my-data.dat")
```

Details of files associated with deposits are stored in a `data.frame` held as
`cli$hostdata$files` which after running that command will include a row with
an item whose "filename" will be "my-data.dat".

```{r uploaded-files, eval = FALSE}
cli$hostdata$files
#>                           checksum    filename filesize
#> 1 5955bb96a8fee3bc89549bde9ef9b470 my-data.dat      829
#>                                     id
#> 1 618ae9b9-af48-4b86-aa37-7b4e767dccb7
#>                                                 links.download
#> 1 https://sandbox.zenodo.org/api/files/<file-hash>/my-data.dat
#>                                                                links.self
#> 1 https://sandbox.zenodo.org/api/deposit/depositions/1065666/files/<hash>
```


The `deposit_download_file()`
function does the reverse:

```{r deposit_download, eval = FALSE}
path <- cli$deposit_download_file ("my-data.dat")
```

Files are by default downloaded to the current working directory, or elsewhere
specified by an additional `path` parameter. The `deposit_download_file()`
function is the only main client function which returns a value, in this case
of the full path to the downloaded file.

### Additional functions

A deposits client offers a few functions beyond those demonstrated above.
The command `ls(cli)` lists all elements of the R6 client, including all
variables and functions, with functions prefixed with `deposit_` or
`deposits_`. Additional functions include:


1. `deposit_delete(deposit_id)` to delete a nominated deposit;
2. `deposit_retrieve(deposit_id)` to populate a local deposits client with
   information on a specified deposit;
3. `deposit_update(deposit_id)` to update deposit on server with local
   metadata;
4. `deposits_list()` to update lists of current deposits within client.
5. `deposits_search()` to search deposits in an online service.

See [the introductory
vignette](https://docs.ropensci.org/deposits/articles/deposits.html) for more
details.

## Code of Conduct

Please note that this package is released with a [Contributor Code of
Conduct](https://ropensci.org/code-of-conduct/). By contributing to this
project, you agree to abide by its terms.
