---
title: "deposits"
output:
  md_document:
    variant: gfm

  rmarkdown::html_vignette:
    self_contained: no
---

# deposits

<!-- badges: start -->
[![R build status](https://github.com/ropenscilabs/deposits/workflows/R-CMD-check/badge.svg)](https://github.com/ropenscilabs/deposits/actions?query=workflow%3AR-CMD-check)
[![codecov](https://codecov.io/gh/ropenscilabs/deposits/branch/main/graph/badge.svg)](https://codecov.io/gh/ropenscilabs/deposits)
[![Project Status: Concept](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
<!-- badges: end -->


The `deposits` R package is a universal client for depositing and accessing
research data anywhere. Currently supported services are
[zenodo](https://zenodo.org) and [figshare](https://figshare.com).

## Installation

The package can be installed by enabling the ["ropensci"
r-universe](https://ropensci.r-universe.dev), and using `install.packages()`:

```{r r-univ, eval = FALSE}
options (repos = c (
    ropensci = 'https://ropensci.r-universe.dev',
    CRAN = 'https://cloud.r-project.org'))
install.packages ("deposits")
```

Alternatively, the package can be installed directly from GitHub with the
following command:

```{r remotes-install, eval = FALSE}
remotes::install_github ("ropenscilabs/deposits")
```

The package can then be loaded for use with:

```{r}
library (deposits)
```

## Data Repositories

The list of data repositories currently supported is accessible by the
`deposits_services()` function:

```{r services}
deposits_services ()
```

[`zenodo`](https://zenodo.org) offers a "sandbox" environment, which
offers an ideal environment for testing the functionality of this
package.


## API tokens

All services require users to create an account and then to generate API
("Application Programming Interface") tokens. Click on the following links to
generate tokens, also listed as sequences of menu items used to reach token
settings:

- [zenodo/account/settings/applications/tokens/new](https://zenodo.org/account/settings/applications/tokens/new/)
- [zenodo-sandbox/account/settings/applications/tokens/new](https://sandbox.zenodo.org/account/settings/applications/tokens/new/),
- [figshare/account/applications](https://figshare.com/account/applications).

It is not necessary to create or register applications on any of these
services; this package uses personal tokens only. The tokens need to be 
stored as local environment variables the names of which must include the names
of the respective services, as defined by the "name" column returned from
`deposits_service()`, as shown above. This can be done as in the following
example:

```{r key, eval = FALSE}
Sys.setenv ("ZENODO_SANDBOX_TOKEN" = "<my-token")
```

Alternatively, these tokens can be stored in a `~/.Renviron` file, where they
will be automatically loaded into every R session.

## Functionality

The `deposits` package extends from [an `R6`
client](https://github.com/r-lib/R6) offering a variety of methods for
performing actions on deposits services. Details of the methods can be seen in
the [help file for the
`depositsClient`](https://docs.ropensci.org/deposits/reference/depositsClient.html).
All `deposits` operations start with a client constructed with the `new()`
function, with a first parameter specifying the desired deposits service:

```{r client-new}
cli <- depositsClient$new ("zenodo")
cli
cli <- depositsClient$new ("zenodo", sandbox = TRUE)
cli
```

A `deposits` client offers the following functions, listed here with names of
parameters:

1. `ping()` to verify Authorization tokens;
2. `list_deposits()` to do just that;
3. `fill_metadata(meta)` to fill a `deposits` client with metadata (see
   Metadata vignette for details);
4. `new_deposit()` to create a new deposit using the metadata inserted with
   `fill_metadata()`;
5. `delete_deposit(deposit_id)` to delete a nominated deposit;
6. `upload_file(deposit_id, path)` to upload a local file to a specified deposit;
7. `retrieve_deposit(deposit_id)` to retrieve metadata for a nominated deposit; and
8. `download_file(deposit_id, filename)` to download nominated `filename` from
   specified deposit.

[An introductory
vignette](https://docs.ropensci.org/deposits/articles/deposits.html)
demonstrates all of these functions, while the "Metadata" vignette describes
processes to specify deposit metadata with a `deposits` client.

## Code of Conduct

Please note that this package is released with a [Contributor Code of
Conduct](https://ropensci.org/code-of-conduct/). By contributing to this
project, you agree to abide by its terms.
