---
title: "deposits"
output:
  md_document:
    variant: gfm

  rmarkdown::html_vignette:
    self_contained: no
---

# deposits

<!-- badges: start -->
[![R build status](https://github.com/ropenscilabs/deposits/workflows/R-CMD-check/badge.svg)](https://github.com/ropenscilabs/deposits/actions?query=workflow%3AR-CMD-check)
[![codecov](https://codecov.io/gh/ropenscilabs/deposits/branch/main/graph/badge.svg)](https://codecov.io/gh/ropenscilabs/deposits)
[![Project Status: Concept](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
<!-- badges: end -->


The `deposits` R package is a universal client for depositing and accessing
research data anywhere. Currently supported services are
[zenodo](https://zenodo.org) and [figshare](https://figshare.com). The package
works seamlessly with [the "frictionless" data
workflow](https://frictionlessdata.io/). Instructions for installing and
setting up the package are in [a separate
vignette](https://docs.ropensci.org/deposits/articles/install-setup.html). This
README gives a brief overview of package usage, with more detail in [the
introductory
vignette](https://docs.ropensci.org/deposits/articles/deposits.html).


## Data Repositories

The list of data repositories currently supported is accessible by the
`deposits_services()` function:

```{r services}
library (deposits)
deposits_services ()
```

The [`zenodo`](https://zenodo.org) "sandbox" environment offers an ideal
environment for testing the functionality of this package.


## The "frictionless" data workflow

Prior to describing the `deposits` workflow, it is useful to understand the
["frictionless" data workflow](https://frictionlessdata.io/), which provides a
standardised way to document the contents and structure of data files through a
separate metadata file. The `deposits` package links directly with [rOpenSci's
`frictionless` package](https://docs.ropensci.org/frictionless) for writing,
editing, and reading frictionless metadata files.

The `deposits` package nevertheless aims to make this integration as seamless
and painless as possible. No knowledge of the frictionless workflow is
necessary, and the `deposits` workflow should "just work" in many cases,
appropriately constructing frictionless metadata files and uploading them to
deposits services. It is of course also possible to use the `deposits` package
without frictionless metadata.

## The deposits workflow

The `deposits` package uses [the `R6` package](https://github.com/r-lib/R6) to
create [a
`depositsClient`](https://docs.ropensci.org/deposits/reference/depositsClient.html)
used to perform all functions. All `deposits` operations start with a client
constructed with the `new()` function:

```{r client-new, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>    hostdata : <none>
#>    metadata : <none>
```

The upper section of information shown by printing the client describes
information on the client in general, and aspects of the nominated service
associated with the user (identified through the token, as described in [the
"Installation and Setup"
vignette](https://docs.ropensci.org/deposits/articles/install-setup.html#setup-api-tokens)).
The lower section contains information on the current deposit held within the
client. A new client initially contains no information on a deposit, and so
these lower sections are empty.

The code above demonstrates how `R6` objects work, through calling "methods" or
functions on the main "object" via the `$` symbol. The "new" method must be
called on [a `depositsClient`
object](https://docs.ropensci.org/deposits/reference/depositsClient.html). From
that point on, all other methods are called on that object itself. For example,
the following calls lists all methods implemented by the client:

```{r deposits-methods, eval = FALSE}
cli$deposits_methods ()
#> List of methods for a deposits client:
#>
#>    - deposit_delete
#>    - deposit_download_file
#>    - deposit_fill_metadata
#>    - deposit_new
#>    - deposit_retrieve
#>    - deposit_service
#>    - deposit_update
#>    - deposit_upload_file
#>    - deposits_list
#>    - deposits_methods
#>    - deposits_search
#>
#>  see `?depositsClient` for full details of all methods.
```

All of those methods then work similarly, by calling
`cli$<name_of_method>(<parameters>)`. A typical deposits workflow, and
associated methods, involves the following steps:

1. Fill the client with metadata (using the `new()` or
   `deposit_fill_metadata()` methods);
2. Create a new deposit on the nominated service (using the `deposit_new()`
   method); and
3. Upload files for the deposit (using the `deposit_upload_file()` method).

### Fill client with metadata

The `deposits` package ultimately aims to simplify the management of data on
external repositories. All of these platforms require additional metadata
describing the data to be deposited, with the `deposits` package working with
two main types of metadata:

1. Metadata which describe a deposit and associated properties, such as author
   names and affiliations, deposit titles and descriptions, dates, keywords, 
   links to other deposits or publications, and many other terms.
2. Metadata which describe the actual contents of the data to be deposited.
   These kinds of metadata are (optionally) generated and (always) handled here
   by [the `frictionless`
   package](https://docs.ropensci.org/frictionless).

As explained at the outset, the `deposits` package can be used without knowing
or understanding anything about [the `frictionless`
package](https://docs.ropensci.org/frictionless) or [frictionless
workflows](https://frictionlessdata.io). Most of this present section therefore
concerns the first of these two kinds of metadata, with brief demonstrations of
the second kind used to illustrate how the frictionless workflow integrates
with the general `deposits` workflow. The term "metadata" throughout the
following should be understood to refer to the first of the above kinds of
metadata; the second kind is always referred to here as "frictionless
metadata."

Metadata can be included in a deposits client in two main ways, either on
initial construction, or through attaching metadata to an existing client. In
both cases, metadata may be specified in a variety of ways, one of the easiest
of which is to specify metadata is as a list of named terms like the following
example:

```{r metadata}
metadata <- list (
    title = "New Title",
    abstract = "This is the abstract",
    creator = list ("A. Person", "B. Person")
)
```

These data can be used in construction of a new client by passing a `metadata`
argument:

```{r client-new-metadata, eval = FALSE}
cli <- depositsClient$new (
    service = "zenodo",
    sandbox = TRUE,
    metadata = metadata
)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>    hostdata : <none>
#>    metadata : 4 terms (see 'metadata' element for details)
```

The "metadata" component of the client now holds 4 terms (including an
additional "created" datestamp). Equivalently, metadata can be added to an
existing client with [the `deposit_fill_metadata()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_fill_metadata):

```{r client-fill-metadata, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
cli$deposit_fill_metadata (metadata)
```

Other ways of specifying and entering metadata are described in [the
introductory
vignette](https://docs.ropensci.org/deposits/articles/deposits.html).

Note that R6 functions are called directly on the client, with the object
itself (`cli`) updated by the call. There is thus no need to call,
```{r client-fill-return, eval = FALSE}
cli <- cli$deposit_fill_metadata (metadata)
```
as the `cli` object is updated by the call as shown above, and does not need to
be assigned to any return value.

### Create a new deposit

The metadata filled with the above steps can then be used to initiate a new
deposit on the associated server using the `deposit_new()` function:

```{r deposit_new, eval = FALSE}
cli$deposit_new ()
#> ID of new deposit : 1065666
print (cil)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>  url_service : https://sandbox.zenodo.org/deposit/1065666
#>   deposit id : 1065666
#>     hostdata : list with 14  elements
#>     metadata : 4 terms (see 'metadata' element for details)
```

The client now includes several additional elements, notably a "deposit id"
(stored in `cli$id`) giving the unique identifier for the new deposit, and a
`hostdata` item with, in this case,  14 elements as specified by the host
service. The `url_service` is the URL for the newly-created deposit. (Viewing
in a web browser will require logging in for all private and sandbox deposits).

### Uploading (and downloading) files

A deposit is really about data, not just metadata. Uploading data to a deposit
is as simple as the following function:

```{r deposit_upload, eval = FALSE}
cli$deposit_upload_file (path = "<path>/<to>/my-data.dat")
```

Details of files associated with deposits are stored in a `data.frame` held as
`cli$hostdata$files` which after running that command will include a row with
an item whose "filename" will be "my-data.dat".

```{r uploaded-files, eval = FALSE}
cli$hostdata$files
#>                           checksum    filename filesize
#> 1 5955bb96a8fee3bc89549bde9ef9b470 my-data.dat      829
#>                                     id
#> 1 618ae9b9-af48-4b86-aa37-7b4e767dccb7
#>                                                 links.download
#> 1 https://sandbox.zenodo.org/api/files/<file-hash>/my-data.dat
#>                                                                links.self
#> 1 https://sandbox.zenodo.org/api/deposit/depositions/1065666/files/<hash>
```


The `deposit_download_file()` function does the reverse:

```{r deposit_download, eval = FALSE}
path <- cli$deposit_download_file ("my-data.dat")
```

Files are by default downloaded to the current working directory, or elsewhere
specified by an additional `path` parameter. The `deposit_download_file()`
function is the only main client function which returns a value, in this case
of the full path to the downloaded file.

### Additional functions

A deposits client offers a few functions beyond those demonstrated above.
The command `ls(cli)` lists all elements of the R6 client, including all
variables and functions, with functions prefixed with `deposit_` or
`deposits_`. Additional functions include:


1. `deposit_delete(deposit_id)` to delete a nominated deposit;
2. `deposit_retrieve(deposit_id)` to populate a local deposits client with
   information on a specified deposit;
3. `deposit_update(deposit_id)` to update deposit on server with local
   metadata;
4. `deposits_list()` to update lists of current deposits within client.
5. `deposits_search()` to search deposits in an online service.

See [the introductory
vignette](https://docs.ropensci.org/deposits/articles/deposits.html) for more
details.

## Code of Conduct

Please note that this package is released with a [Contributor Code of
Conduct](https://ropensci.org/code-of-conduct/). By contributing to this
project, you agree to abide by its terms.
