---
title: "deposits"
output:
  md_document:
    variant: gfm

  rmarkdown::html_vignette:
    self_contained: no
---

# deposits

<!-- badges: start -->
[![R build status](https://github.com/ropenscilabs/deposits/workflows/R-CMD-check/badge.svg)](https://github.com/ropenscilabs/deposits/actions?query=workflow%3AR-CMD-check)
[![codecov](https://codecov.io/gh/ropenscilabs/deposits/branch/main/graph/badge.svg)](https://codecov.io/gh/ropenscilabs/deposits)
[![Project Status: Concept](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
<!-- badges: end -->


The `deposits` R package is a universal client for depositing and accessing
research data anywhere. Currently supported services are
[zenodo](https://zenodo.org) and [figshare](https://figshare.com). The package
works seamlessly with [the "frictionless" data
workflow](https://frictionlessdata.io/). Instructions for installing and
setting up the package are in [a separate
vignette](https://docs.ropensci.org/deposits/articles/install-setup.html). This
README gives a brief overview of package usage, with more detail in [the
introductory
vignette](https://docs.ropensci.org/deposits/articles/deposits.html). For those
who prefer to jump straight in to a workflow, there is also [a deposits
workflow](https://docs.ropensci.org/deposits/articles/example-workflow.html)
vignette demonstrating a full workflow from data generation and documentation
to publication.


## Data Repositories

The list of data repositories currently supported is accessible by the
`deposits_services()` function:

```{r services}
library (deposits)
deposits_services ()
```

The [`zenodo`](https://zenodo.org) "sandbox" environment offers an ideal
environment for testing the functionality of this package.


## The "frictionless" data workflow

Prior to describing the `deposits` workflow, it is useful to understand the
["frictionless" data workflow](https://frictionlessdata.io/), which provides a
standardised way to document the contents and structure of data files through a
separate metadata file. The `deposits` package links directly with [rOpenSci's
`frictionless` package](https://docs.ropensci.org/frictionless) for writing,
editing, and reading frictionless metadata files.

The `deposits` package nevertheless aims to make this integration as seamless
and painless as possible. No knowledge of the frictionless workflow is
necessary, and the `deposits` workflow should "just work" in many cases,
appropriately constructing frictionless metadata files and uploading them to
deposits services. It is of course also possible to use the `deposits` package
without frictionless metadata.

## The deposits workflow

This section gives a brief overview of the deposits workflow. A more complete
description is given in the [main package
vignette](https://docs.ropensci.org/deposits/articles/deposits.html), and
demonstrated step-by-step in [the workflow
vignette](https://docs.ropensci.org/deposits/articles/example-workflow.html).
The `deposits` package uses [the `R6` package](https://github.com/r-lib/R6) to
create [a
`depositsClient`](https://docs.ropensci.org/deposits/reference/depositsClient.html)
used to perform all functions. All `deposits` operations start with a client
constructed with the `new()` function:

```{r client-new, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>    hostdata : <none>
#>    metadata : <none>
```

The upper section of information shown by printing the client describes general
information, and aspects of the specified service, and the profile of the user
(identified through the locally-stored token, as described in [the
"Installation and Setup"
vignette](https://docs.ropensci.org/deposits/articles/install-setup.html#setup-api-tokens)).
The lower section contains information on the current deposit held within the
client. A new client initially contains no information on a deposit, and so
these lower sections are empty.

The code above demonstrates how `R6` objects work, through calling "methods" or
functions on the main "object" via the `$` symbol. The "new" method must be
called on [a `depositsClient`
object](https://docs.ropensci.org/deposits/reference/depositsClient.html). From
that point on, all other methods are called on that object itself. For example,
the following calls lists all methods implemented by the client:

```{r deposits-methods, eval = FALSE}
cli$deposits_methods ()
#> List of methods for a deposits client:
#>
#>    - deposit_add_resource
#>    - deposit_delete
#>    - deposit_delete_file
#>    - deposit_download_file
#>    - deposit_embargo
#>    - deposit_fill_metadata
#>    - deposit_new
#>    - deposit_prereserve_doi
#>    - deposit_publish
#>    - deposit_retrieve
#>    - deposit_service
#>    - deposit_update
#>    - deposit_upload_file
#>    - deposits_list
#>    - deposits_methods
#>    - deposits_search
#>
#>  see `?depositsClient` for full details of all methods.
```

All of those methods then work similarly, by calling
`cli$<method>(<parameters>)`. Note that R6 functions are called directly on the
client, with the object itself (`cli`) updated by the call. For example, [the
`deposit_fill_metadata()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-fill-metadata-)
described below can be used to specify metadata for a deposit. This works by
simply calling,

```{r client-fill-return, eval = FALSE}
cli$deposit_fill_metadata (metadata)
```
The `cli` object is then updated by this call, without needing to be assigned
to a return value (like `cli <- cli$<method>`).

A typical deposits workflow, and associated methods, involves the following
steps:

1. Fill the client with metadata (using the
   [`new()`](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-new-)
   or
   [`deposit_fill_metadata()`](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_fill_metadata) methods);
2. Create a new deposit on the nominated service (using the
   [`deposit_new()`](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-new-)
   method); and
3. Upload files for the deposit (using the
   [`deposit_upload_file()`](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-upload-file-)
   method).

The following sections describe each of these steps in more detail.

### Fill client with metadata

The *metadata* of a deposit describe the nature and properties of the data
being deposited. A deposit is first created by uploading metadata to a deposits
service; only then can actual data be uploaded and associated with the deposit
described by the metadata. The `deposits` package works with two main types of
metadata:

1. Metadata which describe a deposit and associated properties, such as author
   names and affiliations, deposit titles and descriptions, dates, keywords, 
   links to other deposits or publications, and many other terms.
2. Metadata which describe the actual contents of the data to be deposited.
   These kinds of metadata are (optionally) generated and (always) handled here
   by [the `frictionless`
   package](https://docs.ropensci.org/frictionless).

As explained at the outset, the `deposits` package can be used without knowing
or understanding anything about [the `frictionless`
package](https://docs.ropensci.org/frictionless) or [frictionless
workflows](https://frictionlessdata.io). Most of this present section therefore
concerns the first of these two kinds of metadata, referred to throughout all
package documentation as "metadata". Brief demonstrations are also given of the
second kind, to illustrate how the frictionless workflow integrates with the
general `deposits` workflow. These kinds of metadata are always referred to as
"frictionless metadata."

A separate vignette describes in detail [how to specify and include metadata in
a `deposits`
client](https://docs.ropensci.org/deposits/articles/metadata.html). This
section briefly demonstrates the process. An example of `deposits` metadata is:

```{r metadata}
metadata <- list (
    title = "New Title",
    abstract = "This is the abstract",
    creator = list (list (name = "A. Person"), list (name = "B. Person"))
)
```

The "creator" field is a list-of-lists, to allow individual creator entries to
have multiple fields in addition to "name". These data can be used in
construction of a new client by passing a `metadata` argument:

```{r client-new-metadata, eval = FALSE}
cli <- depositsClient$new (
    service = "zenodo",
    sandbox = TRUE,
    metadata = metadata
)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>    hostdata : <none>
#>    metadata : 3 terms (see 'metadata' element for details)
```

The "metadata" component of the client now holds 3 terms. Equivalently,
metadata can be added to an existing client with [the `deposit_fill_metadata()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_fill_metadata):

```{r client-fill-metadata, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
cli$deposit_fill_metadata (metadata)
```

At that stage, the metadata are only associated with the local client. The
following section describes how to use those metadata to initiate a deposit on
an external service.


### Create a new deposit

The metadata filled with the above steps can then be used to initiate a new
deposit on the associated server using [the `deposit_new()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-new-).

```{r deposit_new, eval = FALSE}
cli$deposit_new ()
#> ID of new deposit : 1065666
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>  url_service : https://sandbox.zenodo.org/deposit/1065666
#>   deposit id : 1065666
#>     hostdata : list with 14  elements
#>     metadata : 4 terms (see 'metadata' element for details)
```

The client now includes several additional elements, notably a "deposit id"
(stored in `cli$id`) giving the unique identifier for the new deposit, and a
"hostdata" item with, in this case,  14 elements as specified by the host
service. The `url_service` is the URL for the newly-created deposit. (Viewing
in a web browser will require logging in for all private and sandbox deposits).
The "metadata" item also includes an additional "identifier" element containing
a pre-reserved DOI provided by the deposits service.

### Uploading (and downloading) files

A deposit is really about data, not just metadata. Data can be uploaded to a deposit
with [the `deposit_upload_file()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-upload-file-):

```{r deposit_upload, eval = FALSE}
cli$deposit_upload_file (path = "<path>/<to>/my-data.dat")
```

Details of files associated with deposits are stored in a `data.frame` held as
`cli$hostdata$files` which after running that command will include a row with
an item whose "filename" will be "my-data.dat".

```{r uploaded-files, eval = FALSE}
cli$hostdata$files
#>                           checksum    filename filesize
#> 1 5955bb96a8fee3bc89549bde9ef9b470 my-data.dat      829
#>                                     id
#> 1 618ae9b9-af48-4b86-aa37-7b4e767dccb7
#>                                                 links.download
#> 1 https://sandbox.zenodo.org/api/files/<file-hash>/my-data.dat
#>                                                                links.self
#> 1 https://sandbox.zenodo.org/api/deposit/depositions/1065666/files/<hash>
```


The `deposit_download_file()` function does the reverse:

```{r deposit_download, eval = FALSE}
path <- cli$deposit_download_file ("my-data.dat")
```

Files are by default downloaded to the current working directory, or elsewhere
specified by an additional `path` parameter. Unlike most methods for deposit
clients, which invisibly return an updated version of the client, the
`deposit_download_file()` returns the full local path to the downloaded file.

## Code of Conduct

Please note that this package is released with a [Contributor Code of
Conduct](https://ropensci.org/code-of-conduct/). By contributing to this
project, you agree to abide by its terms.

## Contributors



<!-- ALL-CONTRIBUTORS-LIST:START - Do not remove or modify this section -->
<!-- prettier-ignore-start -->
<!-- markdownlint-disable -->

All contributions to this project are gratefully acknowledged using the [`allcontributors` package](https://github.com/ropenscilabs/allcontributors) following the [all-contributors](https://allcontributors.org) specification. Contributions of any kind are welcome!

### Code

<table>

<tr>
<td align="center">
<a href="https://github.com/mpadge">
<img src="https://avatars.githubusercontent.com/u/6697851?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/deposits/commits?author=mpadge">mpadge</a>
</td>
</tr>

</table>


### Issues

<table>

<tr>
<td align="center">
<a href="https://github.com/collinschwantes">
<img src="https://avatars.githubusercontent.com/u/6107885?u=414bad1b3dd9499b3dc60180fc411f1fb18de24e&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/deposits/issues?q=is%3Aissue+author%3Acollinschwantes">collinschwantes</a>
</td>
<td align="center">
<a href="https://github.com/Bisaloo">
<img src="https://avatars.githubusercontent.com/u/10783929?u=38e3754466eaa200e20f0609709467b6331cdfbe&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/deposits/issues?q=is%3Aissue+author%3ABisaloo">Bisaloo</a>
</td>
<td align="center">
<a href="https://github.com/maelle">
<img src="https://avatars.githubusercontent.com/u/8360597?u=824f03caa87c92420352e3dd9a05470320a67412&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/deposits/issues?q=is%3Aissue+author%3Amaelle">maelle</a>
</td>
<td align="center">
<a href="https://github.com/joelnitta">
<img src="https://avatars.githubusercontent.com/u/13459362?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/deposits/issues?q=is%3Aissue+author%3Ajoelnitta">joelnitta</a>
</td>
<td align="center">
<a href="https://github.com/noamross">
<img src="https://avatars.githubusercontent.com/u/571752?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/deposits/issues?q=is%3Aissue+author%3Anoamross">noamross</a>
</td>
<td align="center">
<a href="https://github.com/peterdesmet">
<img src="https://avatars.githubusercontent.com/u/600993?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropenscilabs/deposits/issues?q=is%3Aissue+author%3Apeterdesmet">peterdesmet</a>
</td>
</tr>

</table>

<!-- markdownlint-enable -->
<!-- prettier-ignore-end -->
<!-- ALL-CONTRIBUTORS-LIST:END -->
