---
title: "The deposits R6 client"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{2. The deposits R6 client}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

The deposits package uses [an `R6` client](https://github.com/r-lib/R6) to
interface with the individual deposition services. This vignette briefly
explains explains the properties of a deposits client for those unfamiliar
with `R6` objects.

### R6 methods

The `R6` package used to construct deposits clients here allows for
structured class objects in R. The objects include elements (such as variables)
and methods, which for R are generally functions. A new client can be
constructed with the `new` operator, which for deposits requires specifying
the service for which the client is to be constructed:

```{r new-cli, eval = FALSE}
library (deposits)
cli <- depositsClient$new (service = "figshare")
```

Additional functions are called in a similar way, using a `$`-notation, in the
form `cli$deposit_function()`. The deposits package is constructed so that
function calls constructed is this way will "automatically" update the object
itself, and so generally do not need to be assigned to a return value. For
example, the function `deposits_list()` updates the list of deposits on the
associated service. In conventional R packages, calling this function would
require assigning a return value like this:

```{r cli-list-deps, eval = FALSE}
cli_updated <- cli$deposits_list ()
```

`R6` objects are, however, always updated internally, so the client itself,
`cli`, will already include the updated list of deposits without any need for
assigning the return value to `cli_updated`. That is, rather than the above
line, all deposits functions may be called simply as,

```{r cli-list-deps2, eval = FALSE}
cli$deposits_list ()
```

(The single exception to this is the `deposit_download_file()` function, which
returns the path to the locally downloaded file, and so should always be
assigned to a return value.)


### Initialising a deposits client

An empty client can be constructed by naming the desired service. An additional
`sandbox` parameter constructs a client to the `zenodo` sandbox environment
intended for testing their API. Actual use of the `zenodo` API can then be
enabled with the default `sandbox = FALSE`.

```{r client-new, eval = FALSE}
cli <- depositsClient$new ("zenodo", sandbox = TRUE)
cli
#> <deposits client>
#> deposits service : zenodo
#>           sandbox: TRUE
#>         url_base : https://sandbox.zenodo.org/api/
#> Current deposits : <none>
#>
#>   hostdata : <none>
#>   metadata : <none>
```

### The R6 client structure

All `R6` clients share a few build-in methods which can be used to understand
their internal structure and functionality. In particular, both [the
`ls()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/ls.html) and
[str()
function](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/str.html)s
reveal internal details of R6 clients. These details are nevertheless not
particularly helpful here, and the deposits client has it's own
[`deposits_methods()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposits-methods-)
intended to provide an overview all all implemented methods:

```{r methods, eval = FALSE}
cli$deposits_methods ()
#> List of methods for a deposits client:
#>
#>    - deposit_add_resource
#>    - deposit_delete
#>    - deposit_delete_file
#>    - deposit_download_file
#>    - deposit_embargo
#>    - deposit_fill_metadata
#>    - deposit_new
#>    - deposit_prereserve_doi
#>    - deposit_publish
#>    - deposit_retrieve
#>    - deposit_service
#>    - deposit_update
#>    - deposit_upload_file
#>    - deposits_list
#>    - deposits_methods
#>    - deposits_search
#>
#>  see `?depositsClient` for full details of all methods.
```

As described in the final line of that output, the documentation contains full
details of all methods, also provided in [the online
documentation](https://docs.ropensci.org/deposits/reference/depositsClient.html#methods).
