---
title: "The deposits package"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{The deposits package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

The `deposits` package is a universal client for depositing and accessing
research data anywhere. Currently supported services are
[zenodo](https://zenodo.org) and [figshare](https://figshare.com).
These two systems have fundamentally different interfaces ("API"s, or
Application Programming Interfaces), and access to these and
indeed all deposition services has traditionally been enabled through
individual software clients. The `deposits` package aims to be a
universal client offering access to a variety of deposition services, without
users having to know any specific details of the APIs for each service. This
vignette provides a demonstration of the primary functionality of the
`deposits` package.

```{r lib}
library (deposits)
library (atom4R) # for metadata objects
```


## The deposits client

The `deposits` package uses [an `R6` client](https://github.com/r-lib/R6) to
interface with the individual deposition services. An empty client can be
constructed by naming the desired service, where the `sandbox` parameter
constructs a client to the `zenodo` sandbox environment intended for testing
their API. Actual use of the `zenodo` API can then be enabled with the default
`sandbox = FALSE`.

```{r client-new, eval = FALSE}
cli <- depositsClient$new ("zenodo", sandbox = TRUE)
cli
#> <deposits client>
#>     name: zenodo
#>  sandbox: TRUE
#>     url : https://sandbox.zenodo.org/api/
#> metadata: <none>
```

Client construction requires local environment variables to be stored with
personal access keys, as described in the [main `README`
document](https://docs.ropensci.org/deposits/#api-keys). Clients can be tested
with the `ping()` function, to confirm authorization, and the `list_deposits()`
function can be used to list all of your personal deposits.

```{r ping-list, eval = FALSE}
cli$ping ()
#> [1] TRUE
deps <- cli$list_deposits ()
class (deps)
#> [1] "data.frame"
names (deps)
#>  [1] "conceptrecid" "created"      "id"           "links"        "metadata"
#'  [6] "modified"     "owner"        "record_id"    "state"        "submitted"
#' [11] "title"        "doi"          "doi_url"
```

The result of `list_deposits()` differ for the two services, but both return
`data.frame` objects. Importantly, these `data.frame`s all have a column named
`id` which holds unique, integer-valued identifiers for the deposits. These
identifiers are required for several of `deposit` functions, in which they are
passed as a `deposit_id` parameter.

## Metadata

The primary function of `deposits` is `new_deposit()`, which creates a new
deposit using metadata passed to the `deposits` client. The specification and
construction of metadata is described in detail in a second vignette, with this
vignette providing a brief overview. `deposits` metadata is represented as a
[`DCEntry` object from the `atom4R`
package](https://github.com/eblondel/atom4R/wiki#atom4R-create-dublincore).
The "DC" stands for the ["Dublin Core Metadata Initiative"
(DCMI)](https://www.dublincore.org/) which provides arguably the most
widely-used [set of metadata terms and
entities](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/).
The `DCEntry` objects are also `R6` objects, constructed by default with no
metadata:

```{r dcentry-demo}
library (atom4R)
meta <- DCEntry$new ()
meta
```

The metadata in a `deposits` object can be defined either on construction of a
new object, or using the `fill_metadata()` function:

```{r deposit-metadata-fill, eval = FALSE}
cli <- depositsClient$new ("zenodo", sandbox = TRUE, metadata = meta)
cli
#> <deposits client>
#>     name: zenodo
#>  sandbox: TRUE
#>     url : https://sandbox.zenodo.org/api/
#> metadata:<DCEntry>
#> ....|-- updated: 2022-02-23 15:48:11
# or:
cli <- depositsClient$new ("zenodo", sandbox = TRUE)
cli$fill_metadata (meta)
```

### Filling metadata

The documentation for [the `atom4R`
package](https://github.com/eblondel/atom4R/wiki#atom4R-create-dublincore)
describes in detail how to fill metadata in a `DCEntry()` object. An example
is:

```{r dc-meta-fill, eval = FALSE}
meta$addDCTitle ("My new deposit")
cli
#> <deposits client>
#>     name: zenodo
#>  sandbox: TRUE
#>     url : https://sandbox.zenodo.org/api/
#> metadata:<DCEntry>
#> ....|-- updated: 2022-02-23 15:48:11
#> ....|-- title <DCTitle>
#> ........|-- value: My new deposit
```

Note that as `R6` objects, both the `meta` and `cli` objects are pointers,
which enables the update to the `meta` object in the first of the lines above
to be immediately reflected in an updated `deposits` object. To make metadata
completion easier, the `deposits` package enables metadata to be entered in a
simple text file using the following workflow:

1. Run [the `deposits_metadata_template()`
   function](https://docs.ropensci.org/deposits/reference/deposits_metadata_template.html)
   specifying a local file which will be written with an empty metadata
   template in a simple JSON format;
2. Edit the template and complete any desired metadata; and
3. Run [the `deposits_meta_to_dcmi()`
   function](https://docs.ropensci.org/deposits/reference/deposits_meta_to_dcmi.html)
   to use the file to automatically populate a `DCEntry` object.

The following code demonstrates how to create and modify a metadata template,
including a line to define the single "title" metadata entry.

```{r meta-demo}
f <- tempfile (fileext = ".json")
deposits_metadata_template (filename = f)
template <- readLines (f)
head (template)
title_line <- grep ("Title", template)
template [title_line]
template [title_line] <- "  \"Title\": \"My new deposit\","
writeLines (template, f)
```

Note that all quotation marks are "escaped" - that is, prefixed with
backslashes, and that each line ends with a comma. Construction of a `deposits`
metadata object is then as simple as the following line:

```{r meta-to-dc}
meta <- deposits_meta_to_dcmi (f)
```

That metadata can then be used to construct a `deposits` client, in which the
title entered in the template then appears as part of the `DCEntry` metadata
object.

```{r new-client, eval = FALSE}
cli <- depositsClient$new ("zenodo", sandbox = TRUE, metadata = meta)
cli
#> <deposits client>
#>     name: zenodo
#>  sandbox: TRUE
#>     url : https://sandbox.zenodo.org/api/
#> metadata:<DCEntry>
#> ....|-- updated: 2022-02-23 15:48:11
#> ....|-- title <DCTitle>
#> ........|-- value: My new deposit
```

## Creating and deleting new deposits

The following single line demonstrates how to create a new deposit:

```{r new_deposit, eval = FALSE}
res <- cli$new_deposit ()
class (res)
#> [1] "list"
names (res)
#>  [1] "conceptrecid" "created"      "doi"          "doi_url"      "files"
#>  [6] "id"           "links"        "metadata"     "modified"     "owner"
#> [11] "record_id"    "state"        "submitted"    "title"
res$title
#> [1] "My new deposit"
res$id
#> [1] 1021945
```

The list of deposits should then include the newly constructed one:

```{r chedk-dep-id, eval = FALSE}
deps <- cli$list_deposits ()
res$id %in% deps$id
#> [1] TRUE
```

Deposits can be deleted by passing `id` values to `delete_deposit()`:

```{r, eval = FALSE}
cli$delete_deposit (res$id)
#> [1] TRUE
```

## Uploading files to deposits

Deposits created with the `new_deposit()` function contain metadata without any
associated files, such as binary data files. These can be uploaded with a
deposit using the `upload_file()` function, which accepts two parameters of
`deposit_id`, and a `path` variable specifying a local file to be uploaded.
The following lines create an arbitrary binary file to be uploaded.

```{r bin-file, eval = FALSE}
path <- tempfile ()
dat <- paste0 (sample (c (LETTERS, letters, 0:9), size = 1e5, replace = TRUE),
               collapse = "")
con <- file (path, open = "wb")
writeBin (dat, con)
close (con)
```

Uploading to our deposit is then as simple as:

```{r upload-demo, eval = FALSE}
d <- cli$new_deposit ()
res <- cli$upload_file (deposit_id = d$id, path = path)
class (res)
#> [1] "list"
names (res)
#> [1] "mimetype"      "updated"       "links"         "is_head"       "created"
#> [6] "checksum"      "version_id"    "delete_marker" "key"           "size"
```

The `deposit_id` can then be used to get details of the deposit, and to confirm
that our newly uploaded file has been included.

```{r upload-check-fake-data, echo = FALSE}
links <- data.frame (
download = "https://sandbox.zenodo.org/api/files/a619f5d0-034c-4cb1-b03c-7d01984d0e76/file13934345cf839",
self = "https://sandbox.zenodo.org/api/deposit/depositions/1022056/files/1989dff8-69bd-46cb-81c8-0b8d28b5bfef"
)

files <- data.frame (
    checksum = "01352a8608e77c6d5a85e18238baf88f",
    filename = "file13934345cf839",
    filesize = 100194,
    id = "1989dff8-69bd-46cb-81c8-0b8d28b5bfef",
    links = links)

res <- list (
    "conceptrecid" = "",
    "created" = "",
    "doi" = "",
    "doi_url" = "",
    "files"= files,
    "id" = "",
    "links" = "",
    "metadata" = "",
    "modified" = "",
    "owner" = "",
    "record_id" = "",
    "state" = "",
    "submitted" = "",
    "title" = "")
```

```{r upload-check, eval = FALSE}
res <- cli$retrieve_deposit (deposit_id = d$id)
names (res)
#>  [1] "conceptrecid" "created"      "doi"          "doi_url"      "files"
#>  [6] "id"           "links"        "metadata"     "modified"     "owner"
#> [11] "record_id"    "state"        "submitted"    "title"
```
The "files" component has the details of the file:
```{r upload-file-details}
res$files
```
