---
title: "The deposits package"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{The deposits package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

The `deposits` package is a universal client for depositing and accessing
research data anywhere. Currently supported services are
[zenodo](https://zenodo.org) and [figshare](https://figshare.com).
These two systems have fundamentally different interfaces ("API"s, or
Application Programming Interfaces), and access to these and
indeed all deposition services has traditionally been enabled through
individual software clients. The `deposits` package aims to be a
universal client offering access to a variety of deposition services, without
users having to know any specific details of the APIs for each service. This
vignette provides a demonstration of the primary functionality of the
`deposits` package.

```{r lib}
library (deposits)
```


## The deposits client

The `deposits` package uses [an `R6` client](https://github.com/r-lib/R6) to
interface with the individual deposition services. An empty client can be
constructed by naming the desired service, where the `sandbox` parameter
constructs a client to the `zenodo` sandbox environment intended for testing
their API. Actual use of the `zenodo` API can then be enabled with the default
`sandbox = FALSE`.

```{r client-new}
cli <- depositsClient$new ("zenodo", sandbox = TRUE)
cli
```

Client construction requires local environment variables to be stored with
personal access keys, as described in the [main `README`
document](https://docs.ropensci.org/deposits/#api-keys). Clients can be tested
with the `ping()` function, to confirm authorization, and the `list_deposits()`
function can be used to list all of your personal deposits.

```{r ping-list, eval = FALSE}
cli$ping ()
#> [1] TRUE
deps <- cli$list_deposits ()
class (deps)
#> [1] "data.frame"
names (deps)
#>  [1] "conceptrecid" "created"      "id"           "links"        "metadata"
#'  [6] "modified"     "owner"        "record_id"    "state"        "submitted"
#' [11] "title"        "doi"          "doi_url"
```

The result of `list_deposits()` differ for the two services, but both return
`data.frame` objects. Importantly, these `data.frame`s all have a column named
`id` which holds unique, integer-valued identifiers for the deposits. These
identifiers are required for several of `deposit` functions, in which they are
passed as a `deposit_id` parameter.

## Metadata

The primary function of `deposits` is `new_deposit()`, which creates a new
deposit using metadata passed to the `deposits` client. The specification and
construction of metadata is described in detail in a second vignette, with this
vignette providing a brief overview. `deposits` metadata is represented as a
[`DCEntry` object from the `atom4R`
package](https://github.com/eblondel/atom4R/wiki#atom4R-create-dublincore).
The "DC" stands for the ["Dublin Core Metadata Initiative"
(DCMI)](https://www.dublincore.org/) which provides arguably the most
widely-used [set of metadata terms and
entities](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/).
The `DCEntry` objects are also `R6` objects, constructed by default with no
metadata:

```{r}
library (atom4R)
meta <- DCEntry$new ()
meta
```

The metadata in a `deposits` object can be defined either on construction of a
new object, or using the `fill_metadata()` function:

```{r deposit-metadata-fill}
cli <- depositsClient$new ("zenodo", sandbox = TRUE, metadata = meta)
cli
# or:
cli <- depositsClient$new ("zenodo", sandbox = TRUE)
cli$fill_metadata (meta)
```

### Filling metadata

The documentation for [the `atom4R`
package](https://github.com/eblondel/atom4R/wiki#atom4R-create-dublincore)
describes in detail how to fill metadata in a `DCEntry()` object. An example
is:

```{r dc-meta-fill}
meta$addDCTitle ("My new deposit")
cli
```

Note that as `R6` objects, both the `meta` and `cli` objects are pointers,
which enables the update to the `meta` object in the first of the lines above
to be immediately reflected in an updated `deposits` object. To make metadata
completion easier, the `deposits` package enables metadata to be entered in a
simple text file using the following workflow:

1. Run [the `deposits_metadata_template()`
   function](https://docs.ropensci.org/deposits/reference/deposits_metadata_template.html)
   specifying a local file which will be written with an empty metadata
   template in a simple JSON format;
2. Edit the template and complete any desired metadata; and
3. Run [the `deposits_meta_to_dcmi()`
   function](https://docs.ropensci.org/deposits/reference/deposits_meta_to_dcmi.html)
   to use the file to automatically populate a `DCEntry` object.

The following code demonstrates how to create and modify a metadata template,
including a line to define the single "title" metadata entry.

```{r meta-demo}
f <- tempfile (fileext = ".json")
deposits_metadata_template (filename = f)
template <- readLines (f)
head (template)
title_line <- grep ("Title", template)
template [title_line]
template [title_line] <- "  \"Title\": \"My new deposit\","
writeLines (template, f)
```

Note that all quotation marks are "escaped" - that is, prefixed with
backslashes, and that each line ends with a comma. Construction of a `deposits`
metadata object is then as simple as the following line:

```{r}
meta <- deposits_meta_to_dcmi (f)
```

That metadata can then be used to construct a `deposits` client, in which the
title entered in the template then appears as part of the `DCEntry` metadata
object.

```{r}
cli <- depositsClient$new ("zenodo", sandbox = TRUE, metadata = meta)
cli
```
