---
title: "The deposits package"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{The deposits package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

"deposits" is an R package which provides a universal client for depositing and
accessing research data in a variety of online deposition services. Currently
supported services are [zenodo](https://zenodo.org) and
[figshare](https://figshare.com). These two systems have fundamentally
different interfaces ("API"s, or Application Programming Interfaces), and
access to these and indeed all deposition services has traditionally been
enabled through individual software clients. The deposits package aims to be a
universal client offering access to a variety of deposition services, without
users having to know any specific details of the APIs for each service. This
vignette demonstrates how the deposits package can be used to manage the
processes of uploading and publishing research data, using the methods
summarised in Figure 1.

```{r lib, echo = FALSE}
library (deposits)
```

![Figure 1: Schematic diagram of deposits methods, each of which is explained below. All methods have prefixes of "deposit_", which are not shown in the diagram. Methods connecting to the central "deposits Client" all modify its internal properties, whereas the "upload_file" and "download_file" methods copy files between local storage and remote services without modifying the deposits client itself. The methods connecting the client to "Local Storage" only have an effect the first time they are called, when they create a local version of a frictionless "datapackage.json" file, if one did not previously exist.](fig1.png)

## The deposits client

The deposits package uses [an `R6` client](https://github.com/r-lib/R6) to
interface with the individual deposition services. A [separate
vignette](https://docs.ropensci.org/deposits/articles/deposits-R6.html)
describes the `R6` system for those unfamiliar with it.

### Initialising a deposits client

An empty client can be constructed by naming the desired service. An additional
`sandbox` parameter constructs a client to the `zenodo` sandbox environment
intended for testing their API. Actual use of the `zenodo` API can then be
enabled with the default `sandbox = FALSE`.

```{r client-new, eval = FALSE}
cli <- depositsClient$new ("zenodo", sandbox = TRUE)
cli
#> <deposits client>
#> deposits service : zenodo
#>           sandbox: TRUE
#>         url_base : https://sandbox.zenodo.org/api/
#> Current deposits : <none>
#>
#>   hostdata : <none>
#>   metadata : <none>
```

Client construction requires personal access or authentication tokens for
deposits services to be stored as local environment variables, as described in
the [installation and setup
document](https://docs.ropensci.org/deposits/articles/install-setup.html#setup-api-tokens).
Authentication tokens are checked when new clients are constructed, so the
`$new()` function will only succeed with valid tokens.

As also described [in the
`README`](https://docs.ropensci.org/deposits/index.html#the-deposits-workflow),
all methods of a deposits client can be seen with [the `deposits_methods()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposits-methods-):

```{r ls-cli, eval = FALSE}
cli$deposits_methods ()
#> List of methods for a deposits client:
#>
#>    - deposit_add_resource
#>    - deposit_delete
#>    - deposit_delete_file
#>    - deposit_download_file
#>    - deposit_embargo
#>    - deposit_fill_metadata
#>    - deposit_new
#>    - deposit_prereserve_doi
#>    - deposit_publish
#>    - deposit_retrieve
#>    - deposit_service
#>    - deposit_update
#>    - deposit_upload_file
#>    - deposits_list
#>    - deposits_methods
#>    - deposits_search
#>
#>  see `?depositsClient` for full details of all methods.
```

```{r count-methods, echo = FALSE}
# x <- capture.output (cli$deposits_methods ())
# num_methods <- length (grep ("\\-\\sdeposits\\_", x))
num_methods <- 3L
```


All methods are described in detail in the [documentation entry for the
deposits
client](https://docs.ropensci.org/deposits/reference/depositsClient.html#methods).
All methods starting with the singular "deposit_" prefix operate on individual
deposits. The final `r num_methods` methods starting with "deposits_" are
general methods applied to services in general ("list" and "search"), or to a
deposits client in general ("methods"). The main methods, and relationships
between them, are also illustrated in Figure 1.

## Metadata

The client constructed above is mostly empty, but nevertheless demonstrates the
two primary fields or elements of a deposits client, the *"hostdata"* and
*"metadata"*. Both of these elements represent the "metadata" of a deposit,
with the data itself referred to as "files", which can be uploaded and
downloaded. These files also have accompanying metadata, according to [the
"frictionless" workflow](https://frictionlessdata.io/) as described in the
[separate "frictionless"
vignette](https://docs.ropensci.org/deposits/articles/frictionless.html).

There are thus three types of metadata used in a deposits workflow:

1. *"metadata"* which describe a deposit and associated properties, such as author
   names and affiliations, deposit titles and descriptions, dates, keywords, 
   links to other deposits or publications, and many other terms. These kinds
   of metadata are described in [the metadata
   vignette](https://docs.ropensci.org/deposits/articles/metadata.html).
2. *"frictionless metadata"* which describe the actual contents of the data to
   be deposited. These kinds of metadata are (optionally) generated and
   (always) handled here by [the `frictionless`
   package](https://docs.ropensci.org/frictionless). These kind of metadata are
   described in [the frictionless
   vignette](https://docs.ropensci.org/deposits/articles/frictionless.html).
3. *"hostdata"* which are provided in different formats by the various
   deposits services, and are intended as *read-only* data used to examine
   the remote records of a deposit.

The term "metadata" refers through this and all deposits documentation to the
first of these three kinds, with the second always explicitly referred to as
"frictionless metadata." The "metadata" and "frictionless metadata" structures
remain consistent between services, and allow data to be transformed from one
format to another, and between local clients and remote services. In contrast,
the `hostdata` structures are directly provided by the deposits host services,
generally as lists, and with different structures for different services. These
structures are read-only fields which are automatically filled by the deposits
client, and are intended to provide insight into metadata records stored on
host sites.


### Passing metadata to a deposits client

A new deposit is initially constructed by filling the `metadata` field with a
local representation of metadata. There are several ways of doing this, as
described in [the separate metadata
vignette](https://docs.ropensci.org/deposits/articles/metadata.html). One of
the easiest approaches is to define metadata as a simple list:

```{r metadata-example-1, eval = TRUE}
metadata <- list (
    title = "New Title",
    abstract = "This is the abstract",
    creator = list (list (name = "A. Person"), list (name = "B. Person"))
)
```

Note that the "creator" item has to be a list-of-lists, because aspects other
than name may also be included, and the second list is required to distinguish
different creators, as described in detail in [the metadata
vignette](https://docs.ropensci.org/deposits/articles/metadata.html). A new
deposits client can be filled with this metadata by passing it as the
`metadata` parameter:

```{r new-cli-with-meta-1, eval = FALSE}
cli <- depositsClient$new (
    service = "zenodo",
    sandbox = TRUE,
    metadata = metadata
)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>     hostdata : <none>
#>     metadata : 3 terms (see 'metadata' element for details)
```

The summary produced by calling `print()` (or, equivalently, just typing `cli`
in the console) says that the object now includes three metadata terms. They
can be seen by viewing `cli$metadata`, confirming that the client metadata are
precisely what we specified:

```{r cli-dcmi-metadata, eval = FALSE}
cli$metadata
```
```{r cli-metadata-fakey, echo = FALSE}
metadata <- list (
    abstract = "This is the abstract",
    creator = list (list (name = "A. Person"), list (name = "B. Person")),
    title = "New Title"
)
print (metadata)
```

Alternative ways of specifying and entering metadata are described in [the
metadata vignette](https://docs.ropensci.org/deposits/articles/metadata.html),
along with detailed descriptions of the kinds of metadata accepted by a
deposits client.


## Creating a new deposit


Once filled with metadata, a deposits client can be used to initiate a new
deposit on the associated external service with [the `$deposit_new()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-new-).
The [`$deposit_new()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-new-)
uses an existing client to create a new deposit on the nominated service,
whereas the [the `$new()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-new-)
method creates a new client. Calling `deposit_new()` from the client
constructed above with our sample metadata gives the following result:

```{r deposit_new, eval = FALSE}
cli$deposit_new ()
#> ID of new deposit: 1064327
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : 1 (see 'deposits' element for details)
#>
#>  url_deposit : https://sandbox.zenodo.org/deposit/1064327
#>   deposit id : 1064327
#>     hostdata : list with 14  elements
#>     metadata : 4 terms (see 'metadata' element for details)
```

The client now lists one current deposit, additional fields for the URL and
"id" of the deposit, and has a "hostdata" field with 14 elements. The "ID"
value printed by the call to `deposit_new()` is listed in the client as its
"deposit id". This is a unique integer value used to identify particular
deposits on external services. The value can be accessed any time as `cli$id`.
The "metadata" item also includes an additional "identifier" element containing
a pre-reserved DOI provided by the deposits service.

From that point on, a client will always show (at least) one deposit. For
example, if we return at some later time to a new R session and initiate a new,
empty client, we would see the following result:

```{r new-cli-after-deposit, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : 1 (see 'deposits' element for details)
#>
#>   hostdata : <none>
#>   metadata : <none>
```

This differs from our initial client in that it now lists one "current
deposit".

## Retrieving deposits

We can examine a deposits client to get the "id" values of all current
deposits. Extending from the previous example, the "id" can be accessed as:

```{r id-value, eval = FALSE}
cli$deposits$id
#> [1] 1064327
```

More generally, information of all deposits currently associated with a user's
account (as identified by the token described in the [installation
vignette](https://docs.ropensci.org/deposits/articles/install-setup.html#setup-api-tokens))
can be accessed as `cli$deposits`. With the single deposit show in the previous
steps, the first few fields of the result look this this:

```{r deposits-list, eval = FALSE}
cli$deposits [, 1:5]
```
```{r deposits-list-output, echo = FALSE}
data.frame (
    conceptrecid = 1200932L,
    created = "2023-00-01T00:00:00",
    doi = "10.5072/zenodo.1064327",
    doi_url = "https://doi.org/10.5072/zenodo.1064327",
    id = 1064327L
)
```

We can retrieve the metadata from this or any previously uploaded deposit with
the `deposit_retrieve()` function:

```{r deposit-retrieve, eval = FALSE}
cli$deposit_retrieve (deposit_id = cli$deposits$id [1])
```

The local client then holds identical information to the previous client
immediately after calling `deposit_new()` - that is, `retrieve_deposit()` has
filled the local client with all of the metadata from the previously-created
deposit.


## Uploading files to deposits

The previous sections of this document describe how to initiate a deposits
client, and how to use that to initiate and retrieve metadata from a remote
deposits services. The main point of a deposit is of course to store actual
data in any arbitrary format alongside these structured metadata. This is
achieved with [the `deposit_upload_file()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_upload_file),
demonstrated in the following code which uses our deposit retrieved directly
above. It is recommended to store all data for a single deposit within a single
directory, which the following code also creates.


```{r upload-file, eval = FALSE}
data_dir <- file.path (tempdir (), "data")
dir.create (data_dir)
path <- file.path (data_dir, "data.csv")
write.csv (datasets::Orange, path, row.names = FALSE)
cli$deposit_upload_file (path = path)
#> frictionless metadata file has been generated as '/tmp/RtmpxSiYhW/data/datapackage.json'
```

Although the `print` output of our `cli` object does not change after
uploading, the details of the files are contained in the `hostdata$files`
element (which was previously empty):

```{r cli-hostdata-files, eval = FALSE}
cli$hostdata$files
#>                           checksum         filename filesize                                   id
#> 1 cc624d72ede85ef061afa494d9951f6f         data.csv      625 56c44dd6-5f84-4212-9a65-d37f64ca886f
#> 2 eaeb7c4f8a931c99e662172299a0b17f datapackage.json      812 32d556ef-5b65-4b9d-a8a8-2e7bed11da5d
#>                                                                               links.download
#> 1         https://sandbox.zenodo.org/api/files/561f4971-9e86-4235-b574-f5662f6088e3/data.csv
#> 2 https://sandbox.zenodo.org/api/files/561f4971-9e86-4235-b574-f5662f6088e3/datapackage.json
#>                                                                                              links.self
#> 1 https://sandbox.zenodo.org/api/deposit/depositions/1161632/files/56c44dd6-5f84-4212-9a65-d37f64ca886f
#> 2 https://sandbox.zenodo.org/api/deposit/depositions/1161632/files/32d556ef-5b65-4b9d-a8a8-2e7bed11da5d
```

The list of files includes a "datapackage.json" file generated by [the
`frictionless` package](https://docs.ropensci.org/frictionless). As described
in [the main README](https://docs.ropensci.org/deposits/index.html), and at
length in [the separate "frictionless"
vignette](https://docs.ropensci.org/deposits/articles/frictionless.html), these
files contain both the metadata entered in to the deposits client, as well as
"frictionless metadata" describing the internal properties of the dataset
itself.

Files can be downloaded with the [`deposit_download_file`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-download-file-).
To demonstrate how that works, the following code first removes the local
version, then downloads it from the remote service and confirms that a local
version has been successfully re-created.

```{r dl_file, eval = FALSE}
file.remove (path)
file <- cli$deposit_download_file (filename = "data.csv", path = data_dir)
file
#> [1] /tmp/RtmpcO59N8/data/data.csv
```

### Pre-generating frictionless metadata files prior to upload

The workflow described in the preceding section results in a frictionless
metadata file being simultaneously generated, filled with deposits metadata,
and uploaded to the nominated service.
As described in detail in [the "frictionless"
vignette](https://docs.ropensci.org/deposits/articles/frictionless.html). An
alternative workflow allows frictionless metadata files to be generated locally
prior to any uploading. This uses [the `deposits_add_resource()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-add-resource-),
where a "resource" is a local data file or object.

After initiating a client with metadata, as demonstrated above:

```{r add-resource-new, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE, metadata = metadata)
```

A frictionless metadata file which is only stored locally can then be generated
by the following call, by specifying a path to that local file.

```{r add-resource, eval = FALSE}
cli$deposit_add_resource (path = path)
```

The directory containing the specified file will then have an additional
"datapackage.json" file including the deposits metadata used in client
construction. This file may be edited as desired prior to uploading. To update
a deposits client with changes to external metadata files, simply pass the path
to that file to [the `deposits_fill_metadata()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-fill-metadata-).
When ready, a single call to [the `deposit_upload_file()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_upload_file)
will upload the file specified in that call, along with the frictionless
"datapackage.json" metadata file.

## Editing and updating deposits

All deposits are initiated on the nominated services as "private" deposits, meaning:

1. They can only be viewed by the deposit owner; and
2. They can be freely edited, including complete deletion.

A deposit can only be publicly viewed once it has been published, as described
in the final section of this vignette. The process of using deposits to
prepare one or more datasets for publication will generally involve multiple
stages of editing and updating. As described in the final section of [the
metadata
vignette](https://docs.ropensci.org/deposits/articles/metadata.html#metadata-and-the-deposits-frictionless-workflow),
the local frictionless metadata file can be edited as desired, and the
corresponding deposit updated both locally and on the remote deposits service
with [the `deposit_fill_metadata()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-fill-metadata-):

```{r metadata-fill-update, eval = FALSE}
cli <- depositsClient$new (service = "<service>")
path <- "/<local>/<directory>/<or>/<frictionless_file>"
cli$deposit_fill_metadata (metadata = path)
```

Alternatively, the `metadata` parameter can be passed directly to the `new()`
method to initiate a new client, as demonstrated immediately below.

## Example workflow

An example of a full workflow for creating and editing a deposits client and
associated metadata would look something like the following five main steps:

1. Initiate local deposits client with metadata:

    ```{r workflow-example-1, eval = FALSE}
    metadata <- list (
        title = "New Title",
        abstract = "This is the abstract",
        creator = list (list (name = "A. Person"), list (name = "B. Person"))
    )
    cli <- depositsClient$new (
        service = "zenodo",
        sandbox = TRUE,
        metadata = metadata
    )
    cli$deposit_new ()
    #> ID of new deposit: 1064327
    ```
2. Upload local data, which the following code simulates by creating a "dummy"
   dataset in the temporary directory of the current R session:
    ```{r local-data-create, eval = FALSE}
    data_dir <- file.path (tempdir (), "data")
    dir.create (data_dir)
    path <- file.path (data_dir, "data.csv")
    write.csv (datasets::Orange, path, row.names = FALSE)
        ```
        The following call then uploads that dataset to the newly-created deposit:
        ```{r local-data-upload, eval = FALSE}
    cli$deposit_upload_file (path = path)
    ```
    Calling `deposit_upload_file()` the first time also creates local and
    remote versions of a frictionless "datapackage.json" file, holding all
    metadata, and the DOI of the new deposit.
3. Modify metadata. The following code provides a proof-of-principle
   modification of metadata, by changing "New Title" to "Updated Title":

    ```{r metadata-modify, eval = FALSE}
    fr <- file.path (data_dir, "datapackage.json")
    dp <- frictionless::read_package (fr)
    dp$metadata$title
    #> [1] "New Title"
    dp$metadata$title <- "Updated Title"
    frictionless::write_package (dp, data_dir)
    ```
    This is an indirect way of editing metadata, by using R code. The
    recommended way to update deposits metadata is to directly edit and modify
    the "datapackage.json" file.
4. Update both local client and remote deposit data:
    ```{r update-metadata, eval = FALSE}
    cli$deposit_update (path = data_dir)
    #> Local file at [/tmp/RtmpBM0VYr/data/data.csv] is identical on host and will not be uploaded.
    #> Local file at [/tmp/RtmpBM0VYr/data/datapackage.json] has changed and will now be uploaded.
    cli$metadata$title
    #> [1] "Updated Title"
    cli$hostdata$title
    #> [1] "Updated Title"
    ```
    Local modificaitons are reflected in both updated "metadata" with the
    deposits client, as well as in "hostdata" stored on the Zenodo service.

## Publishing a deposit

Once all metadata and data have been satisfactorily edited, updated, and
uploaded, a deposit can be made publicly visible and permanently associated
with a Digital Object Identifier (DOI) by publishing it. Prior to publishing,
it is often desired to apply an "embargo" to the deposit, in the form of a date
after which the deposit will become publicly visible. The two steps to
publication are thus generally:

```{r publish, eval = FALSE}
cli$deposit_embargo (embargo_date = "2030-03-30")
cli$deposit_publish ()
```

Calling [the `deposit_publish()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-publish-)
is irreversible, and can never be undone. (Publication is permanent even in the
Zenodo sandbox environment.) The published deposit will be permanently
associated with the account of the user who published it, as identified by [the
API token used to initiate the deposits
client](https://docs.ropensci.org/deposits/articles/install-setup.html#setup-api-tokens).
Publication will also change many items of the client's "hostdata", notably
involving a change of status or visibility from "private" to "public". Once a
deposit has been published, the associated DOI, or equivalent the URL given in
the deposits client, may be shared as a permanent link to the deposit.
