---
title: "The deposits package"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{The deposits package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

The `deposits` package is a universal client for depositing and accessing
research data anywhere. Currently supported services are
[zenodo](https://zenodo.org) and [figshare](https://figshare.com).
These two systems have fundamentally different interfaces ("API"s, or
Application Programming Interfaces), and access to these and
indeed all deposition services has traditionally been enabled through
individual software clients. The `deposits` package aims to be a
universal client offering access to a variety of deposition services, without
users having to know any specific details of the APIs for each service. This
vignette demonstrates how the `deposits` package can be used to manage the
processes of uploading and publishing research data, using the methods
summarised in Figure 1.

```{r lib, echo = FALSE}
library (deposits)
```

![Figure 1: Schematic diagram of 'deposits' methods, each of which is explained below. All methods have prefixes of "deposit_", which are not shown in the diagram. Methods connecting to the central "deposits Client" all modify its internal properties, whereas the "upload_file" and "download_file" methods copy files between local storage and remote services without modifying the deposits client itself. The methods connecting the client to "Local Storage" only have an effect the first time they are called, when they create a local version of a frictionless "datapackage.json" file, if one did not previously exist.](fig1.png)

## The deposits client

The `deposits` package uses [an `R6` client](https://github.com/r-lib/R6) to
interface with the individual deposition services. A [separate
vignette](https://docs.ropensci.org/deposits/articles/deposits-R6.html)
describes the `R6` system for those unfamiliar with it.

### Initialising a deposits client

An empty client can be constructed by naming the desired service. An additional
`sandbox` parameter constructs a client to the `zenodo` sandbox environment
intended for testing their API. Actual use of the `zenodo` API can then be
enabled with the default `sandbox = FALSE`.

```{r client-new, eval = FALSE}
cli <- depositsClient$new ("zenodo", sandbox = TRUE)
cli
#> <deposits client>
#> deposits service : zenodo
#>           sandbox: TRUE
#>         url_base : https://sandbox.zenodo.org/api/
#> Current deposits : <none>
#>
#>   hostdata : <none>
#>   metadata : <none>
```

Client construction requires personal access or authentication tokens for
deposits services to be stored as local environment variables, as described in
the [installation and setup
document](https://docs.ropensci.org/deposits/articles/install-setup.html#setup-api-tokens).
Authentication tokens are checked when new clients are constructed, so the
`$new()` function will only succeed with valid tokens.

As also described [in that
`README`](https://docs.ropensci.org/deposits/index.html#the-deposits-workflow),
all methods of a `deposits` client can be seen with [the `deposits_methods()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposits-methods-):

```{r ls-cli, eval = FALSE}
cli$deposits_methods ()
#> List of methods for a deposits client:
#>
#>    - deposit_add_resource
#>    - deposit_delete
#>    - deposit_delete_file
#>    - deposit_download_file
#>    - deposit_embargo
#>    - deposit_fill_metadata
#>    - deposit_new
#>    - deposit_prereserve_doi
#>    - deposit_publish
#>    - deposit_retrieve
#>    - deposit_service
#>    - deposit_update
#>    - deposit_upload_file
#>    - deposits_list
#>    - deposits_methods
#>    - deposits_search
#>
#>  see `?depositsClient` for full details of all methods.
```

```{r count-methods, echo = FALSE}
# x <- capture.output (cli$deposits_methods ())
# num_methods <- length (grep ("\\-\\sdeposits\\_", x))
num_methods <- 3L
```


Each of these methods is described in detail in the [documentation entry for
the deposits
client](https://docs.ropensci.org/deposits/reference/depositsClient.html#methods).
The final `r num_methods` methods starting with "deposits_" are general methods
applied to services in general ("list" and "search"), or to a deposits client
in general ("methods"). All other methods start with the singular "deposit_"
prefix, and operate on individual deposits. The main methods, and relationships
between them, are also illustrated in Figure 1.

## Metadata

The client constructed above is mostly empty, but nevertheless demonstrates the
two primary fields or elements of a deposits client, the *"hostdata"* and
*"metadata"*. Both of these elements represent the "metadata" of a deposit,
with the data itself referred to as "files", which can be uploaded and
downloaded. These files also have accompanying metadata, according to [the
"frictionless" workflow](https://frictionlessdata.io/) as described in the
[separate "frictionless"
vignette](https://docs.ropensci.org/deposits/articles/frictionless.html).

There are thus three types of metadata used in a `deposits` workflow:

1. *"metadata"* which describe a deposit and associated properties, such as author
   names and affiliations, deposit titles and descriptions, dates, keywords, 
   links to other deposits or publications, and many other terms. These kinds
   of metadata are described in [the metadata
   vignette](https://docs.ropensci.org/deposits/articles/metadata.html).
2. *"frictionless metadata"* which describe the actual contents of the data to
   be deposited. These kinds of metadata are (optionally) generated and
   (always) handled here by [the `frictionless`
   package](https://docs.ropensci.org/frictionless). These kind of metadata are
   described in [the frictionless
   vignette](https://docs.ropensci.org/deposits/articles/frictionless.html).
3. *"hostdata"* which are provided in different formats by the various
   `deposits` services, and are intended as *read-only* data used to examine
   the remote records of a deposit.

The `hostdata` structures are directly provided by the `deposits` host
services, generally as lists, although with different structures for different
services. These structures are read-only fields which are automatically filled
by the `deposits` client, and are intended to provide insight into metadata
records stored on host sites. In contrast, the `metadata` structures remain
consistent between services, and allow data to be transformed from one format
to another, and potentially to allow data to be transferred between different
services.


### Passing metadata to a deposits client

A new deposit is initially constructed by filling the `metadata` field with a
local representation of metadata. There are several ways of doing this, as
described in [the separate metadata
vignette](https://docs.ropensci.org/deposits/articles/metadata.html). One of
the easiest approaches is to define metadata as a simple list:

```{r metadata-example-1, eval = TRUE}
metadata <- list (
    title = "New Title",
    abstract = "This is the abstract",
    creator = list (list (name = "A. Person"), list (name = "B. Person"))
)
```

Note that the "creator" item has to be a list-of-lists, because aspects other
than name may also be included, and the second list is required to distinguish
different creators, as described in detail in [the metadata
vignette](https://docs.ropensci.org/deposits/articles/metadata.html). A new
deposits client can be filled with this metadata by passing it as the
`metadata` parameter:

```{r new-cli-with-meta-1, eval = FALSE}
cli <- depositsClient$new (
    service = "zenodo",
    sandbox = TRUE,
    metadata = metadata
)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>     hostdata : <none>
#>     metadata : 3 terms (see 'metadata' element for details)
```

The summary produced by calling `print()` (or, equivalently, just typing `cli`
in the console) says that the object now includes three metadata terms. They
can be seen by viewing `cli$metadata`, confirming that the client metadata are
precisely what we specified:

```{r cli-dcmi-metadata, eval = FALSE}
cli$metadata
```
```{r cli-metadata-fakey, echo = FALSE}
metadata <- list (
    abstract = "This is the abstract",
    creator = list (list (name = "A. Person"), list (name = "B. Person")),
    title = "New Title"
)
print (metadata)
```

Alternative ways of specifying and entering metadata are described in [the
metadata vignette](https://docs.ropensci.org/deposits/articles/metadata.html),
along with detailed descriptions of the kinds of metadata accepted by a
`deposits` client.


## Creating a new deposit


Once filled with metadata, a `deposits` client can be used to initiate a new
deposit on the associated external service with [the `$deposit_new()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-new-).
This is not to be confused with [the `$new()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-new-)
method creates a new client. The [`$deposit_new()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-new-)
uses an existing client to create a new deposit on the external service. Using
the client constructed above with our sample metadata gives the following
result:

```{r deposit_new, eval = FALSE}
cli$deposit_new ()
#> ID of new deposit: 1064327
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : 1 (see 'deposits' element for details)
#>
#>  url_deposit : https://sandbox.zenodo.org/deposit/1064327
#>   deposit id : 1064327
#>     hostdata : list with 14  elements
#>     metadata : 3 terms (see 'metadata' element for details)
```

The client now lists one current deposit, additional fields for the URL and
"id" of the deposit, and has a `hostdata` field with 14 elements. Importantly,
the `id` field holds a unique integer value used to identify particular
deposits both on all external services, and as the `deposit_id` parameter of
`deposits` client functions.

If we now construct a new, empty client, we see the following result:

```{r new-cli-after-deposit, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : 1 (see 'deposits' element for details)
#>
#>   hostdata : <none>
#>   metadata : <none>
```

This differs from our initial client in that it now lists one "current
deposit". We can examine that to get the associated "id" value:

```{r id-value, eval = FALSE}
cli$deposits$id
#> [1] 1064327
```

We can then retrieve the metadata we previously uploaded with the
`deposit_retrieve()` function:

```{r deposit-retrieve, eval = FALSE}
cli$deposit_retrieve (deposit_id = cli$deposits$deposit_id)
```

The local client then holds identical information to the previous client
immediately after calling `deposit_new()` - that is, `retrieve_deposit()` has
filled the local client with all of the metadata from the previously-created
deposit.


## Uploading files to deposits

The deposits client to that point has only been used to construct and upload
metadata. The main point of a deposit is of course to store actual data in any
arbitrary format alongside these structured metadata. This is achieved with
[the `deposit_upload_file()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_upload_file),
demonstrated in the following code which uses our deposit retrieved directly
above.


```{r upload-file, eval = FALSE}
path <- file.path (tempdir (), "data.csv")
write.csv (datasets::Orange, path, row.names = FALSE)
cli$deposit_upload_file (path = path)
```

Although the `print` output of our `cli` object does not change after
uploading, the details of the files are contained in the `hostdata$files`
element (which was previously empty):

```{r cli-hostdata-files, eval = FALSE}
cli$hostdata$files
#>                           checksum         filename filesize                                   id
#> 1 cc624d72ede85ef061afa494d9951f6f         data.csv      625 56c44dd6-5f84-4212-9a65-d37f64ca886f
#> 2 eaeb7c4f8a931c99e662172299a0b17f datapackage.json      812 32d556ef-5b65-4b9d-a8a8-2e7bed11da5d
#>                                                                               links.download
#> 1         https://sandbox.zenodo.org/api/files/561f4971-9e86-4235-b574-f5662f6088e3/data.csv
#> 2 https://sandbox.zenodo.org/api/files/561f4971-9e86-4235-b574-f5662f6088e3/datapackage.json
#>                                                                                              links.self
#> 1 https://sandbox.zenodo.org/api/deposit/depositions/1161632/files/56c44dd6-5f84-4212-9a65-d37f64ca886f
#> 2 https://sandbox.zenodo.org/api/deposit/depositions/1161632/files/32d556ef-5b65-4b9d-a8a8-2e7bed11da5d
```

The list of files includes a "datapackage.json" file generated by [the
`frictionless` package](https://docs.ropensci.org/frictionless). As described
in [the main README](https://docs.ropensci.org/deposits/index.html), and at
length in [the separate "frictionless"
vignette](https://docs.ropensci.org/deposits/articles/frictionless.html), these
files contain both the metadata entered in to the `deposits` client, as well as
"frictionless metadata" describing the internal properties of the dataset
itself.

Files can be downloaded with the converse `download_file` function,
demonstrated here by first removing the local copy, and then downloading it
from the deposits service:

```{r dl_file, eval = FALSE}
file.remove (path)
file <- cli$deposit_download_file (filename = "data.csv", path = tempdir ())
file
#> [1] /tmp/RtmpcO59N8/data.csv
```

### Pre-generating frictionless metadata files prior to upload

The workflow described in the preceding section results in a frictionless
metadata file being simultaneously generated, filled with `deposits` metadata,
and uploaded to the nominated service.
As described in detail in [the "frictionless"
vignette](https://docs.ropensci.org/deposits/articles/frictionless.html), an
Alternative workflow allows frictionless metadata files to be generated locally
prior to any uploading. This uses [the `deposits_add_resource()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-add-resource-),
where a "resource" is a local data file or object.

After initiating a client with metadata, as demonstrated above:

```{r add-resource-new, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE, metadata = metadata)
```

A local-only frictionless metadata file can then be generated by the following
call, specifying a path to that local file.

```{r add-resource, eval = FALSE}
cli$deposit_add_resource (path = path)
```

The directory containing the specified file will then have an additional
"datapackage.json" file including the `deposits` metadata used in client
construction. This file may be edited as desired prior to uploading. To update
a `deposits` client with changes to external metadata files, simply pass the
path to that file to [the `deposits_fill_metadata()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-fill-metadata-).
When ready, a single call to [the `deposit_upload_file()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_upload_file)
will upload one or more specified files, along with the frictionless
"datapackage.json" metadata file.

## Editing and updating deposits

All deposits are initiated on the nominated services as "private" deposits, meaning:

1. They can only be viewed by the deposit owner; and
2. They can be freely edited, including complete deletion.

A deposit can only be publicly viewed once it has been published, as described
in the final section of this vignette. The process of using `deposits` to
prepare one or more datasets for publication will generally involve multiple
stages of editing and updating. As described in the final section of [the
metadata
vignette](https://docs.ropensci.org/deposits/articles/metadata.html#metadata-and-the-deposits-frictionless-workflow),
the local frictionless metadata file can be edited as desired, and the
corresponding deposit updated both locally and on the remote deposits service
with [the `deposit_fill_metadata()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-fill-metadata-):

```{r metadata-fill-update, eval = FALSE}
cli <- depositsClient$new (service = "<service>")
path <- "/<local>/<directory>/<or>/<frictionless_file>"
cli$deposit_fill_metadata (metadata = path)
```

Alternatively, the `metadata` parameter can be passed directly to the `new()`
method to initiate a new client, as demonstrated immediately below. An example
full workflow for creating and editing a deposits client and associated
metadata would look something like the following five main steps:

1. Initiate local deposits client with metadata:

    ```{r workflow-example-1, eval = FALSE}
    metadata <- list (
        title = "New Title",
        abstract = "This is the abstract",
        creator = list (list (name = "A. Person"), list (name = "B. Person"))
    )
    cli <- depositsClient$new (
        service = "zenodo",
        sandbox = TRUE,
        metadata = metadata
    )
    cli$deposit_new ()
    #> ID of new deposit: 1064327
    ```
2. Upload local data, which the following code simulates by creating a "dummy"
   dataset in the temporary directory of the current R session:
    ```{r local-data-create, eval = FALSE}
    data_dir <- file.path (tempdir (), "data")
    dir.create (data_dir)
    path <- file.path (data_dir, "data.csv")
    write.csv (datasets::Orange, path, row.names = FALSE)
    cli$deposit_upload_file (path = path)
    ```
    The following call then uploads that dataset to the newly-created deposit:
    ```{r local-data-upload, eval = FALSE}
    cli$deposit_upload_file (path = path)
    ```
    That upload also creates local and remote versions of a frictionless
    "datapackage.json" file, holding all metadata, and the DOI of the new
    deposit.
3. Modify metadata. The following code provides a proof-of-principle
   modification of metadata, by changing "New Title" to "Updated Title":

    ```{r metadata-modify, eval = FALSE}
    fr <- file.path (data_dir, "datapackage.json")
    dp <- frictionless::read_package (fr)
    dp$metadata$title
    #> [1] "New Title"
    dp$metadata$title <- "Updated Title"
    frictionless::write_package (dp, data_dir)
    ```
4. Re-initiate deposits client to update local and remote deposits:
    ```{r update-metadata, eval = FALSE}
    cli <- depositsClient$new (
        service = "zenodo",
        sandbox = TRUE,
        metadata = data_dir
    )
    cli$metadata$title
    #> [1] "Updated Title"
    cli$hostdata$title
    #> [1] "Updated Title"
    ```
    And local modificaitons are reflected in both updated "metadata" with the
    deposits client, as well as in "hostdata" stored on the Zenodo service.
5. Versions of data files stored on remote hosts can be updated by uploading
   updated versions of any files:
    ```{r update-data, eval = FALSE
    cli$deposit_upload_file (path = path)
    ```

## Publishing a deposit

Once all metadata and data have been satisfactorily edited, updated, and
uploaded, a deposit can be made publicly visible and permanently associated a
Digital Object Identifier (DOI) by publishing it. Prior to publishing, it is
often desired to apply an "embargo" to the deposit, in the form of a date after
which the deposit will become publicly visible. The two steps to publication
are thus generally:

```{r publish, eval = FALSE}
cli$deposit_embargo (embargo_date = "2030-03-30")
cli$deposit_publish ()
```

Calling [the `deposit_publish()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-publish-)
is irreversible, and can never be undone. The published deposit will be
permanently associated with the account of the user who published it, as
identified by [the API token used to initiate the deposits
client](https://docs.ropensci.org/deposits/articles/install-setup.html#setup-api-tokens).
Publication will also change many items of the client's "hostdata", notably
involving a change of status or visibility from "private" to "public". Once a
deposit has been published, the associated DOI, or equivalent the URL given in
the deposits client, may be shared as a permanent link to the deposit.
