---
title: "The deposits package"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{The deposits package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

The `deposits` package is a universal client for depositing and accessing
research data anywhere. Currently supported services are
[zenodo](https://zenodo.org) and [figshare](https://figshare.com).
These two systems have fundamentally different interfaces ("API"s, or
Application Programming Interfaces), and access to these and
indeed all deposition services has traditionally been enabled through
individual software clients. The `deposits` package aims to be a
universal client offering access to a variety of deposition services, without
users having to know any specific details of the APIs for each service. This
vignette provides a demonstration of the primary functionality of the
`deposits` package.

```{r lib, echo = FALSE}
library (deposits)
```


## The deposits client

The `deposits` package uses [an `R6` client](https://github.com/r-lib/R6) to
interface with the individual deposition services. The following sub-section explains the properties of a `deposits` client for those unfamiliar with `R6` objects.

### R6 methods

The `R6` package used to construct `deposits` clients here allows for
structured class objects in R. The objects include elements (such as variables)
and methods, which for R are generally functions. A new client can be
constructed with the `new` operator, which for `deposits` requires specifying
the service for which the client is to be constructed:

```{r new-cli, eval = FALSE}
cli <- depositsClient$new (service = "figshare")
```

Additional functions are called in a similar way, using the notation,
`cli$deposit_function()`. The `deposits` package is constructed so that
function calls constructed is this way will "automatically" update the object
itself, and so generally do not need to be assigned to a return value.
For example, the function `deposits_list()` updates the list of deposits on the
associated service. In conventional R packages, calling this function would
require assigning a return value like this:

```{r cli-list-deps, eval = FALSE}
cli_updated <- cli$deposits_list ()
```

`R6` objects are, however, always updated internally, so the client itself,
`cli`, will already include the updated list of deposits without any need for
assigning the return value to `cli_updated`. That is, rather than the above
line, all deposits functions may be called simply as,

```{r cli-list-deps2, eval = FALSE}
cli$deposits_list ()
```

(The single exception to this is the `deposit_download_file()` function, which
returns the path to the locally downloaded file, and so should always be
assigned to a return value.)



### Initialising a deposits client

An empty client can be constructed by naming the desired service. An additional
`sandbox` parameter constructs a client to the `zenodo` sandbox environment
intended for testing their API. Actual use of the `zenodo` API can then be
enabled with the default `sandbox = FALSE`.

```{r client-new, eval = FALSE}
cli <- depositsClient$new ("zenodo", sandbox = TRUE)
cli
#> <deposits client>
#> deposits service : zenodo
#>           sandbox: TRUE
#>         url_base : https://sandbox.zenodo.org/api/
#> Current deposits : <none>
#>
#>   hostdata : <none>
#>   metadata : <none>
```

Client construction requires personal access or authentication tokens for
deposits services to be stored as local environment variables, as described in
the [main `README` document](https://docs.ropensci.org/deposits/#api-keys).
Authentication tokens are checked when new clients are constructed, so the
`$new()` function will only succeed with valid tokens.

As also described in that `README`, the main functions of a `deposits` client
are all prefixed with either `deposit_` or `deposits_`, as seen from the result
of `ls(cli)`:

```{r ls-cli, eval = FALSE}
ls (cli)
#>  [1] "deposit_delete"        "deposit_download_file" "deposit_fill_metadata"
#>  [4] "deposit_new"           "deposit_retrieve"      "deposit_service"
#>  [7] "deposit_update"        "deposit_upload_file"   "deposits"
#> [10] "deposits_list"         "deposits_search"       "headers"
#> [13] "hostdata"              "id"                    "initialize"
#> [16] "metadata"              "print"                 "sandbox"
#> [19] "service"               "term_map"              "url_base"
#> [22] "url_service"
```

The client constructed above is mostly empty, but nevertheless demonstrates the
two primary fields or elements of a deposits client:

1. `hostdata` holding all metadata from a "host" or external deposits service
   for a particular deposit; and 
2. `metadata` holding a consistently structured representation of the key
   components of the `hostdata`.

The `hostdata` structures are generally lists, but differ for different
services, whereas the `metadata` structures remain consistent between services,
and allow data to be transformed from one format to another, and, in future
functionality, will allow data to be transferred between different services.

Both of these elements represent the "metadata" of a deposit, with the data
itself referred to as "files", which can be uploaded and downloaded. Thus all
deposits begin with metadata, with the actual data upload only possible once
the initial metadata has been specified and uploaded.


## Metadata

A new deposit is initially constructed by filling the `metadata` field with a
local representation of metadata. The `hostdata` field is filled only after
this initial deposit metadata has been uploaded to the external service. These
two fields thus further differ through `metadata` being a local, and
locally-defined, representation of metadata for a deposit, whereas `hostdata`
are always provided and defined by the external service. The best way to
understand this distinction is through a practical demonstration.

### Metadata as a list

There are several ways of defining metadata for a `deposits` entity, perhaps
the easiest of which is as a simple list:

```{r metadata-example-1, eval = TRUE}
metadata <- list (
    title = "New Title",
    abstract = "This is the abstract",
    creator = list ("A. Person", "B. Person")
)
```

A new deposits client can be filled with this metadata by passing it as the
`metadata` parameter:

```{r new-cli-with-meta-1, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE, metadata = metadata)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>     hostdata : <none>
#>     metadata : 4 terms (see 'metadata' element for details)
```

The summary produced by calling `print()` (or, equivalently, just typing `cli`
in the console) says that the object now includes three metadata terms. They
can be seen by viewing `cli$metadata`:

```{r cli-dcmi-metadata, eval = FALSE}
cli$metadata
```
```{r cli-metadata-fakey, echo = FALSE}
metadata <- list (
    dcmi = list (
        title = "New Title",
        abstract = "This is the abstract",
        creator = list ("A. Person", "B. Person"),
        created = "2022-12-20"
    ),
    service = list (
        created = "2022-12-20",
        metadata = list (
            title = "New Title",
            upload_type = "other",
            creators = list (
                list (name = "A. Person"),
                list (name = "B. Person")
            ),
            description = "This is the abstract"
        )
    )
)
print (metadata)
```


Metadata in `deposits` objects are stored as `DCEntry` objects in a format
provided by [the `atom4R` package](https://github.com/eblondel/atom4R).
These metadata are primarily intended for internal use within the `deposits`
package, and shouldn't generally need to be manipulated directly by users of
this package (although they certainly can be, as illustrated below).

### Metadata from a local file

Another convenient way to specify metadata is to use [the
`deposits_metadata_template()`
funciton](https://docs.ropensci.org/deposits/reference/deposits_metadata_template.html)
to write a local YAML representation of metadata. This function also accepts
the `metadata` parameter of data specified as a list:

```{r}
meta_file <- tempfile (pattern = "meta-", fileext = ".yaml")
deposits_metadata_template (filename = meta_file, metadata = metadata$dcmi)
head (readLines (meta_file))
```

Those metadata can then be directly edited using any text file editor, and the
modified results saved to a local file. The name of that file can then also be
passed as the `metadata` parameter of a new `deposits` client. The following
code thus produces the same results as above:

```{r new-cli-with-meta-file, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE, metadata = meta_file)
```

## Creating a new deposit


A `deposits` client with metadata can be used to initiate a new deposit on the
associated external service with the `$deposit_new()` function. This is not to
be confused with the `$new()` function which creates a new client. The
`$deposit_new()` function uses an existing client to create a new deposit on
the external service. Using the client constructed above with our sample
metadata gives the following result:

```{r deposit_new, eval = FALSE}
cli$deposit_new ()
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : 1 (see 'deposits' element for details)
#>
#>  url_deposit : https://sandbox.zenodo.org/deposit/1064327
#>   deposit id : 1064327
#>     hostdata : list with 14  elements
#>     metadata : 4 terms (see 'metadata' element for details)
```

The client now lists one current deposit, additional fields for the URL and
"id" of the deposit, and has a `hostdata` field with 14 elements. Importantly,
the `id` field holds a unique integer value used to identify particular
deposits both on all external services, and as the `deposit_id` parameter of
`deposits` client functions. The following code demonstrates by constructing a
new client with no metadata.

```{r new-cli-after-deposit, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : 1 (see 'deposits' element for details)
#>
#>   hostdata : <none>
#>   metadata : <none>
```

This differs from our initial client in that it now lists one "current
deposit". We can examine that to get the associated "id" value:

```{r id-value, eval = FALSE}
cli$deposits$id
#> [1] 1064327
deposit_id <- cli$deposits$id
```

We can then retrieve the metadata we previously uploaded with the
`deposit_retrieve()` function:

```{r deposit-retrieve, eval = FALSE}
cli$deposit_retrieve (deposit_id = deposit_id)
```

The local client then holds identical information to the previous client
immediately after calling `deposit_new()` - that is, `retrieve_deposit()` has
filled the local client with all of the metadata from that deposit.


## Uploading files to deposits

The deposits clients thus far have only been used to construct and upload
metadata. The main point of a deposit is of course to store actual data in any
arbitrary format alongside these structured metadata. This is achieved with
[the `deposit_upload_file()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_upload_file),
demonstrated in the following code which uses our deposit retrieved directly
above.


```{r upload-file, eval = FALSE}
path <- file.path (tempdir (), "data.Rds")
saveRDS (datasets::Orange, path)
cli$deposit_upload_file (path = path)
```

Although the `print` output of our `cli` object does not change after
uploading, the details of the files are contained in the `hostdata$files`
element:

```{r cli-hostdata-files, eval = FALSE}
cli$hostdata$files
#>                           checksum               filename filesize
#> 1 22c61006458c44ed979bf3461e330693 data-18e115882cf14.Rds      506
#>                                     id
#> 1 00cad89d-4084-4337-b96a-2f1e232b2b6c
#>                                                                                     links.download
#> 1 https://sandbox.zenodo.org/api/files/e3f727dc-1dfd-4849-88cf-994f630105cf/data-18e115882cf14.Rds
#>                                                                                              links.self
#> 1 https://sandbox.zenodo.org/api/deposit/depositions/1064327/files/00cad89d-4084-4337-b96a-2f1e232b2b6c
```

Files can be downloaded with the converse `download_file` function,
demonstrated here by first removing the local copy, and then downloading it
from the deposits service:

```{r dl_file, eval = FALSE}
file.remove (path)
file <- cli$deposit_download_file (filename = "data.Rds", path = tempdir ())
identical (readRDS (file), datasets::Orange)
#> [1] TRUE
```
