---
title: "Example deposits workflow"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Example deposits workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

This vignette provides an example of a complete deposits workflow, from
initiation of a deposits client through to data publication.

## Initial Metadata

As described in [the metadata
vignette](https://docs.ropensci.org/deposits/articles/metadata.html), deposits
start with metadata describing general aspects of the data being deposited,
such as descriptions, identification of creators, and any other aspects
specified in [the deposits metadata JSON
schema](https://github.com/ropenscilabs/deposits/blob/main/inst/extdata/dc/schema.json).

This workflow will use the same "beaver" datasets as [the metadata
vignette](https://docs.ropensci.org/deposits/articles/metadata.html), from [R's
"datasets"
package](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/beavers.html).
That vignette demonstrated how to use the error messages triggered by
incorrectly specified metadata to work towards the following specification:

```{r}
metadata <- list (
    creator = list (list (name = "P.S. Reynolds")),
    created = "1994-01-01T00:00:00",
    title = "Time-series analyses of beaver body temperatures.",
    description = "Original source of 'beaver' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry.",
    publisher = "John Wiley and Sons",
    isPartOf = list (list (
        identifier = "ark:/13960/t0mt2n370",
        relation = "isPartOf"
    ))
)
```

The "isPartOf" element is a key part of "deposits" metadata, enabling
individual data sources to be systematically related to other resources, in
this case to the book first describing these data which has an associated "ark"
identifier.

Those metadata can then be used to initiate a deposits client with [the `new()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-new),
demonstrated here with `sandbox = TRUE` to use the Zenodo "sandbox" environment.

```{r client-init, eval = FALSE}
cli <- depositsClient$new (
    service = "zenodo",
    sandbox = TRUE,
    metadata = metadata
)
print (cli)
#> <deposits client>
#> deposits service : zenodo
#>           sandbox: TRUE
#>         url_base : https://sandbox.zenodo.org/api/
#> Current deposits : <none>
#>
#>   hostdata : <none>
#>   metadata : 6 terms (see 'metadata' element for details)
```

The metadata can be edited and extended as desired. The metadata recorded in
the deposits client can be updated after each edit with [the
`deposit_fill_metadata()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-fill-metadata-):

```{r fill-meta, eval = FALSE}
cli$deposit_fill_metadata (metadata)
```

An alternative procedure for editing and updating metadata is described in the
following section.

## Preparing data sources

The ["beaver"
data](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/beavers.html)
actually comprises two datasets, "beaver1" and "beaver2", each of which is a
time series of body temperature measurements from an individual beaver. For
data sources to be uploaded by deposits, they must first exist on a local
computer, meaning in this case that copies of these beaver datasets must first
be written to local files.

The deposits package presumes that a single deposits lives within a dedicated
local directory which includes all associated files. Let's start by making a
temporary directory and storing the "beaver" data there:

```{r make-tempdir}
beaver_dir <- file.path (tempdir (), "beaver")
if (!dir.exists (beaver_dir)) {
    dir.create (beaver_dir)
}
bv1 <- file.path (beaver_dir, "beaver1.csv")
write.csv (datasets::beaver1, bv1, row.names = FALSE)
bv2 <- file.path (beaver_dir, "beaver2.csv")
write.csv (datasets::beaver1, bv2, row.names = FALSE)
```

We can then connect the deposits client with that local directory with [the
`deposit_add_resource()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-add-resource-):

```{r add_resource, eval = FALSE}
cli$deposit_add_resource (beaver_dir)
```

Calling that method also writes a ["frictionless" metadata
file](https://docs.ropensci.org/deposits/articles/frictionless.html#local-generation)
to the directory:

```{r fake-dp, echo = FALSE}
write.csv (beaver1, file.path (beaver_dir, "datapackage.json"), row.names = FALSE)
```
```{r}
list.files (beaver_dir)
```

The additional "datapackage.json" file is initially generated by [the
"frictionless" R package](https://docs.ropensci.org/frictionless), which
automatically fills out details of each "resource", or local file, in a
"resources" section. The deposits package then inserts the metadata specified
above into an additional "metadata" section of the file. The following code
illustrates a portion of the contents of that file:

```{r dp-contents, eval = FALSE}
dp <- list.files (beaver_dir, pattern = "datapackage", full.names = TRUE)
readLines (dp) [1:25]
#>  [1] "{"
#>  [2] "  \"profile\": \"tabular-data-package\","
#>  [3] "  \"metadata\": {"
#>  [4] "    \"created\": \"1994-01-01T00:00:00\","
#>  [5] "    \"creator\": ["
#>  [6] "      {"
#>  [7] "        \"name\": \"P.S. Reynolds\""
#>  [8] "      }"
#>  [9] "    ],"
#> [10] "    \"description\": \"Original source of 'beaver' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry.\","
#> [11] "    \"isPartOf\": ["
#> [12] "      {"
#> [13] "        \"identifier\": \"ark:/13960/t0mt2n370\","
#> [14] "        \"relation\": \"isPartOf\""
#> [15] "      }"
#> [16] "    ],"
#> [17] "    \"publisher\": \"John Wiley and Sons\","
#> [18] "    \"title\": \"Time-series analyses of beaver body temperatures.\""
#> [19] "  },"
#> [20] "  \"resources\": ["
#> [21] "    {"
#> [22] "      \"name\": \"beaver1\","
#> [23] "      \"path\": \"beaver1.csv\","
#> [24] "      \"profile\": \"tabular-data-resource\","
#> [25] "      \"format\": \"csv\","
```

Once a frictionless "datapackage.json" file has been generated, any subsequent
editing of metadata should be done by directly editing that file. Editing
should also generally involve extending the automatically-inserted "resource"
metadata describing the structure of the actual files. Edited and updated
versions of metadata can then be loaded into a deposits client like this:

```{r cli-update-meta, eval = FALSE}
cli$deposit_fill_metadata (beaver_dir)
```

Instead of `beaver_dir`, the full path to the local "datapackage.json" file can
also be passed. The following code demonstrates modification and updating of
metadata by first modifying the "title", and then showing that those changes
are reflected in the client itself:

```{r metadata-change, eval = FALSE}
m <- readLines (dp)
i <- grep ("\"title\"", m)
m [i] <- gsub ("Time", "Modified time", m [i])
writeLines (m, dp)

cli$metadata$title # original title
#> [1] "Time-series analyses of beaver body temperatures."
cli$deposit_fill_metadata (beaver_dir)
cli$metadata$title
#> [1] "Modified time-series analyses of beaver body temperatures."
```

In short, metadata editing with deposits is generally done by editing a local
"datapackage.json" file, after which a deposits client can then be updated with
[the `deposit_fill_metadata()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-fill-metadata-).

## Initiating a remote deposit

The metadata held within a deposits client can be used to initiate a remote
deposit on the specified service with [the `deposit_new()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-new-):

```{r dep-new, eval = FALSE}
cli$deposit_new ()
#> ID of new deposit : 1065666
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : <none>
#>
#>  url_service : https://sandbox.zenodo.org/deposit/1065666
#>   deposit id : 1065666
#>     hostdata : list with 14  elements
#>     metadata : 7 terms (see 'metadata' element for details)
```

The client now contains additional "hostdata" elements, containing all data
recorded by Zenodo for that deposit. The default print method for the client
now also lists additional information including a URL for the new deposit, and
a unique identifier.

Metadata can still be edited and updated within a client through the procedures
described above. The metadata held on Zenodo can then be updated by calling
[the `deposit_update()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-update-).
After editing the local "datapackage.json" file, updating thus involves these
two steps:

```{r update-remote, eval = FALSE}
cli$deposit_fill_metadata (beaver_dir)
cli$deposit_update ()
```

## Uploading files to a remote deposit

The main purpose of the deposits package, and of online deposition services, is
to deposit data. This is done by using the deposits client to upload local
files with [the `deposit_upload_file()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-upload-file-).
The main parameter, `path`, can also be either a single file or an entire
directory.

```{r upload-files, eval = FALSE}
cli$deposit_upload_file (beaver_dir)
#> Loading required namespace: frictionless
#> frictionless metadata file has been generated as '/tmp/RtmpCPOaqC/beaver/beaver1.csv'
#> Please make sure you have the right to access data from this Data Package for your intended use.
#> Follow applicable norms or requirements to credit the dataset and its authors.
cli$hostdata$files [, 1:3]
```
```{r hostfiles, echo = FALSE}
hf <- data.frame (
    checksum = c (
        "c8e7ff1e2e4323198b4be5227ff63864",
        "c8e7ff1e2e4323198b4be5227ff63864",
        "4fd4b5167c28a874170ab611daf824e7"
    ),
    filename = c ("beaver1.csv", "beaver2.csv", "datapackage.json"),
    filesize = c (1909, 1909, 1225)
)
print (hf)
```

The "hostdata" of the client now indicate that the three files have been successfully uploaded.
