---
title: "frictionless workflows with deposits"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{frictionless}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

As described in [the main
README](https://docs.ropensci.org/deposits/index.html#the-deposits-workflow),
the `deposits` package is designed to work with [rOpenSci's `frictionless`
package](https://docs.ropensci.org/frictionless) for documentation of datasets.
The [`frictionless` package](https://docs.ropensci.org/frictionless) is an R
implementation of [the general "frictionless"
workflow](https://frictionlessdata.io). To take advantage of the frictionless
workflow in `deposits`, it is only necessary to know that the workflow involves
documenting datasets in a file called
["datapackage.json"](https://frictionlessdata.io/blog/2016/07/21/publish-any/#_2-add-a-datapackage-json-file).

## Generating "datapackage.json" files

The [`frictionless` package](https://docs.ropensci.org/frictionless) can
automatically generate "datapackage.json" files for any tabular input data.
This functionality is included within the `deposits` package to enable the
workflow demonstrated in [the main
vignette](https://docs.ropensci.org/deposits/articles/deposits.html). This
workflow used the following code, starting by creating a local file with some
tabular data in an empty directory.

```{r temp-beaver}
dir.create (file.path (tempdir (), "data"))
path <- file.path (tempdir (), "data", "beaver1.csv")
write.csv (datasets::beaver1, path, row.names = FALSE)
```

We then need to construct a `deposits` client, and specify metadata, in this
case those describing [the `beaver1`
dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/beavers.html).

```{r beaver-metadata, eval = FALSE}
metadata <- list (
    Creator = "P. S. Reynolds",
    Created = "1994-01-01",
    Title = "Time-series analyses of beaver body temperatures"
)
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE, metadata = metadata)
```

These data can then be used to initiate a new deposit on the Zenodo sandbox
service, and to upload the local data.

```{r beaver-new, eval = FALSE}
cli$deposit_new ()
#> ID of new deposit : 1162420
cli$deposit_upload_file (path = path)
```

The client then lists two files that were uploaded, and the local directory
also now has a "datapackage.json" file:

```{r beaver-datapackage, eval = FALSE}
cli$hostdata$files$filename
#> [1] "beaver1.csv"      "datapackage.json"
list.files (file.path (tempdir (), "data"))
#> [1] "beaver1.csv"      "datapackage.json"
```

## Reading and editing "datapackage.json" files

The "datapackage.json" file can be read with the
[`frictionless::read_package()`
function](https://docs.ropensci.org/frictionless/reference/read_package.html),
returning a named list of metadata entries:

```{r frictionless-read, eval = FALSE}
library (frictionless)
path <- file.path (tempdir (), "data", "datapackage.json")
metadata <- read_package (path)
names (metadata)
#> [1] "profile"   "metadata"  "resources" "directory"
```

The "profile", "resources", and "directory" items are all generated by
`frictionless`, while the "metadata" items holds the `deposits` metadata
entered above:

```{r frictionless-metadata, eval = FALSE}
metadata$metadata
#> $created
#> [1] "1994-01-01"
#>
#> $creator
#> [1] "P. S. Reynolds"
#>
#> $title
#> [1] "Time-series analyses of beaver body temperatures"
```

The data in these files can be edited in several ways, primarily through either:

1. Editing the individual list items in R, and saving the result via
   [`frictionless::write_package()`](https://docs.ropensci.org/frictionless/reference/write_package.html);
   or,
2. Directly editing the "datapackage.json" with a text editor.

## Updating "datapackage.json" on deposits service

The remote version of the "datapackage.json" file can be updated by uploading
it with [the `deposit_upload_file()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-upload-file-):

```{r upload-datapackage, eval = FALSE}
cli$deposit_upload_file (path = path)
```
