---
title: "frictionless workflows with deposits"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{frictionless}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

As described in [the main
README](https://docs.ropensci.org/deposits/index.html#the-deposits-workflow),
the `deposits` package is designed to work with [rOpenSci's `frictionless`
package](https://docs.ropensci.org/frictionless) for documentation of datasets.
The [`frictionless` package](https://docs.ropensci.org/frictionless) is an R
implementation of the general ["frictionless"
workflow](https://frictionlessdata.io). All `deposits` metadata, both for a
deposit in general as described in the accompanying [metadata
vignette](https://docs.ropensci.org/deposits/articles/metadata.html), and for
the internal structure of datasets, are stored in "frictionless" metadata files
called
["datapackage.json"](https://frictionlessdata.io/blog/2016/07/21/publish-any/#_2-add-a-datapackage-json-file).

This vignette describes how to generate those files, and how to use them to
manage deposits metadata.

## Generating "datapackage.json" files

The `deposits` package uses the [`frictionless` R
package](https://docs.ropensci.org/frictionless) to automatically generate
frictionless metadata files named "datapackage.json" for any tabular input
data. These files can only be generated for one or more input data sets, and so
can not be generated for `deposits` metadata without accompanying data files or
resources.

The following code is repeated from [the main
vignette](https://docs.ropensci.org/deposits/articles/deposits.html), and
generates a local data file with some tabular data in an empty directory.

```{r temp-beaver, eval = FALSE}
dir.create (file.path (tempdir (), "data"))
path <- file.path (tempdir (), "data", "beaver1.csv")
write.csv (datasets::beaver1, path, row.names = FALSE)
```

We then need to construct a `deposits` client, and specify metadata, in this
case those describing [the `beaver1`
dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/beavers.html).

```{r beaver-metadata, eval = FALSE}
metadata <- list (
    creator = list (list (name = "P. S. Reynolds")),
    created = "1994-01-01T00:00:00",
    title = "Time-series analyses of beaver body temperatures",
    description = "Original source of 'beaver' dataset."
)
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE, metadata = metadata)
```

### Generation by uploading files

The main vignette demonstrates how to use these metadata to initiate a new
deposit on the Zenodo sandbox service, and then to upload the local data.
Calling [the `deposit_upload_file()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-upload-file-)
automatically generates a "datapackage.json" file (if one does not already
exist), adds `deposits` metadata, and also uploads that file to the `deposits`
service.

```{r beaver-new, eval = FALSE}
cli$deposit_new ()
#> ID of new deposit : 1162420
cli$deposit_upload_file (path = path)
```

The client then lists two files that were uploaded, and the local directory
also now has a "datapackage.json" file:

```{r beaver-datapackage, eval = FALSE}
cli$hostdata$files$filename
#> [1] "beaver1.csv"      "datapackage.json"
list.files (file.path (tempdir (), "data"))
#> [1] "beaver1.csv"      "datapackage.json"
```

The `deposits` package inserts an additional "metadata" field into the
frictionless metadata file, containing the `deposits` metadata defined above.
This method of generating frictionless metadata files requires data to first be
uploaded to a service, and will automatically upload the frictionless metadata
file. Any subsequent editing then requires repeated calls to [the
`deposit_upload_file()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-upload-file-)
to update the contents of this file.

### Local generation 

A frictionless metadata file can also first be generated locally, to allow
editing prior to any uploading. This is achieved with [the
`deposit_add_resource()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-add-resource-).
As mentioned, the frictionless workflow requires a data "resource" to exist.
The resource in the above example is the locally-stored "beaver.csv" file.
Presuming this file to exist, we can again initiate a deposit, and then to call
[`deposit_add_resource()`](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-add-resource-),
instead of
[`deposit_upload_file()`](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-upload-file-).

```{r add_resource, eval = FALSE}
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE, metadata = metadata)
cli$deposit_add_resource (path = path)
```

That call will generate a local frictionless metadata file (if one does not
already exist), and fill it with the `deposits` metadata, without initiating a
new deposit or uploading any files. Whether generated and immediately uploading
through calling `deposit_new()`, or locally generated only through calling
`deposit_add_resource()`, the frictionless data file can then be edited and
updated as described in the following sub-section.


## Reading and editing "datapackage.json" files

The "datapackage.json" file can be read with the
[`frictionless::read_package()`
function](https://docs.ropensci.org/frictionless/reference/read_package.html),
returning a named list of metadata entries:

```{r frictionless-read, eval = FALSE}
library (frictionless)
path <- file.path (tempdir (), "data", "datapackage.json")
metadata <- read_package (path)
names (metadata)
#> [1] "profile"   "metadata"  "resources" "directory"
```

The "profile", "resources", and "directory" items are all generated by
`frictionless`, while the "metadata" items holds the `deposits` metadata
entered above:

```{r frictionless-metadata, eval = FALSE}
metadata$metadata
#> $created
#> [1] "1994-01-01T00:00:00"
#>
#> $creator
#> $creator[[1]]
#> $creator[[1]]$name
#> [1] "P. S. Reynolds"
#>
#> $description
#> [1] "Original source of 'beaver' dataset."
#>
#> $title
#> [1] "Time-series analyses of beaver body temperatures"
```

The data in these files can be edited in several ways, primarily through
either:

1. Editing the individual list items in R, and saving the result via
   [`frictionless::write_package()`](https://docs.ropensci.org/frictionless/reference/write_package.html);
   or,
2. Directly editing the "datapackage.json" with a text editor.

The `frictionless` aspects of the metadata are by default automatically
generated by that package, and generally benefit from the kind of editing and
enhancing described in [the main `frictionless`
vignette](https://docs.ropensci.org/frictionless/articles/frictionless.html#create-and-edit-a-data-package),
including such things as adding descriptions of variables. The
`deposits`-specific "metadata" component can also readily be extended and
edited as desired.

## Updating "datapackage.json" on deposits service

The remote version of the "datapackage.json" file can be updated by uploading
it with [the `deposit_upload_file()`
function](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-upload-file-),
passing the full path to that file.

```{r upload-datapackage, eval = FALSE}
cli$deposit_upload_file (path = path, overwrite = TRUE)
```

## Using deposits client to edit frictionless metadata

The metadata in a `deposits` client may be directly edited and updated. The
metadata shown above may, for example, simply be extended like this:

```{r extend-client-metadata, eval = FALSE}
cli$metadata$description <- "A description"
```

The local and remote "datapackage.json" files may be updated to the current
state of a `deposits` client with [the `deposit_update_frictionless()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-depositsClient-deposit_update_frictionless).
This method accepts the optional parameter, `path`, of the directory to local
data and "datapackage.json" files. Calling the method will then update both
local and remote versions of "datapackage.json" to include all changes to
client metadata.

```{r update-frictionless, eval = FALSE}
cli$deposit_update_frictionless (path = dirname (path))
```

Both local and remote versions will then include the additional "description"
in their "metadata" items.
