---
title: "Installation and Setup"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{1. Installation and Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

## Installation

The package can be installed by enabling the ["ropensci"
r-universe](https://ropensci.r-universe.dev), and using `install.packages()`:

```{r r-univ, eval = FALSE}
options (repos = c (
    ropensci = "https://ropensci.r-universe.dev",
    CRAN = "https://cloud.r-project.org"
))
install.packages ("deposits")
```

Alternatively, the package can be installed directly from GitHub with the
following command:

```{r remotes-install, eval = FALSE}
remotes::install_github ("ropenscilabs/deposits")
```

The package can then be loaded for use with:

```{r}
library (deposits)
```

## Setup: API Tokens

All services require users to create an account and then to generate API
("Application Programming Interface") tokens. Click on the following links to
generate tokens, also listed as sequences of menu items used to reach token
settings:

- [zenodo/account/settings/applications/tokens/new](https://zenodo.org/account/settings/applications/tokens/new/)
- [zenodo-sandbox/account/settings/applications/tokens/new](https://sandbox.zenodo.org/account/settings/applications/tokens/new/),
- [figshare/account/applications](https://figshare.com/account/applications).

It is not necessary to create or register applications on any of these
services; this package uses personal tokens only. 

The tokens need to be 
stored as local environment variables the names of which must be

- ZENOODO_TOKEN
- ZENODO_SANDBOX_TOKEN
- FIGSHARE_TOKEN

in which case the tokens are read from these environmental variables. If the respective one (responding
to the `service` and `sandbox` arguments of `deposits_service()`) is not found,
a heuristic algorytm is used to find environmental variables which include the names
of the respective services, as defined by the "name" column returned from
`deposits_service()`, as shown above. This can be done as in the following
example:

```{r key, eval = FALSE}
Sys.setenv ("ZENODO_SANDBOX_TOKEN" = "<my-token")
```

Alternatively, these tokens can be stored in a `~/.Renviron` file, where they
will be automatically loaded into every R session.
