---
title: "Metadata and Data Entities"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Metadata and Data Entities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

This vignette gives a tabular overview of how the two systems currently
implemented ([zenodo](https://zenodo.org) and
[figshare](https://figshare.com/)) align with the official entities defined in
the [Dublin Core Metadata Initiative
(DCMI)](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/).
The names of recognised DCMI entities or "terms" can be seen with:


```{r DC-terms, eval = TRUE}
library (deposits)
dcmi_terms ()
```

The corresponding entities in the zenodo API are defined at
[developers.zenodo.org](https://developers.zenodo.org/), and those for figshare
at [docs.figshare.com](https://docs.figshare.com/).

### [DCMI Terms](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/#section-2)

Zenodo has a separate "Metadata Resource," elements of which are noted "(m)".

DC | Zenodo | Figshare
--- | --- | ---
abstract | description(m) (also DC description) | description
accessRights | access_right(m)/access_conditions(m) | 
accrualMethod | | 
accrualPeriodicity | | 
accrualPolicy | | 
alternative | | 
audience | communities(m) |
available | | 
bibliographicCitation | journal/conference/imprint/partof/thesis metadata(m(m)) | 
conformsTo | | 
created | created | timeline:firstOnline
dateAccepted | publication_date(m)/dates(m) | timeline
dateCopyrighted | publication_date(m)/dates(m) | timeline
dateSubmitted | publication_date(m)/dates(m) | timeline
educationLevel | | 
extent | | 
hasFormat | | 
hasPart | files/related_identifiers(m) | 
hasVersion | version(m) (see also DC isVersionOf/identifier) | 
instructionalMethod | | 
isFormatOf | | 
isPartOf | | projects/collections
isReferencedBy | | projects/collections
isReplacedBy | | 
isRequiredBy | | 
issued | | 
isVersionOf | version(m) (see also DC hasVersion/identifier) | 
license | license(m) | license
mediator | | 
medium | upload_type(m)/image_type(m) | defined_type
modified | modified | 
provenance | | 
references | references(m) | references
replaces | | 
requires | | 
rightsHolder | | 
spatial | locations(m) (see also DC location) | 
tableOfContents | | 
temporal | | 
valid | | 


### [DCMI Element Properties](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/#section-3)

DC | Zenodo | Figshare
--- | --- | ---
contributor | contributors(m) | 
coverage | locations(m) (see also DC spatial) | 
creator | creators(m)/owner | authors
date | dates(m) | timeline
description | description(m) (also DC abstract)/method(m)/state | description
format | | defined_type
identifier | doi(m)/doi_url(m)/record_id/record_url | 
language | language(m) | 
publisher | | 
relation | | 
rights | | 
source | | 
subject | keywords(m)/subjects(m) | categories (per defined list)
title | title(m)/version(m) | 
type | upload_type(m) | 

## Additional Data & Metadata Elements not covered by DCMI

### Zenodo

- embargo_date(m)
- prereserve_doi(m)
- notes(m)
- grants(m)
- submitted

### Figshare

- tags
- keywords
- funding/funding_list
