---
title: "deposits metadata"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{deposits metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    error = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
```

This vignette demonstrates procedures for specifying and validate metadata for
the `deposits` package. This vignette will use metadata for the two "beaver"
datasets provided with R's ["datasets"
package](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/beavers.html).
Relevant metadata for these datasets are given in the "Source" listed on that
page (or `?beaver1` to see a local version):

> P. S. Reynolds (1994) *Time-series analyses of beaver body temperatures.*
> Chapter 11 of Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L.
> and Greenhouse, J. eds (1994) Case Studies in Biometry. New York: John Wiley
> and Sons. 

```{r lib, echo = FALSE}
library (deposits)
```

## Dublin Core Metadata

The metadata structure used by deposits follows the specifications of the
[Dublin Core Metadata
Initiative](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/).
Users do not need to understand or read any documents on the DCMI structure, as
the implementation is explicitly contained within [the package's
"dc/schema.json"
file](https://github.com/ropenscilabs/deposits/blob/main/inst/extdata/dc/schema.json),
which should be the only document needed to understand metadata compliance. The
above link to the DCMI specifications may nevertheless be useful for anyone
wishing to understand individual terms in a broader context.

## Metadata Validation

Metadata in `deposits` are validated against [JSON
schemas](https://json-schema.org), with the schema for the main "metadata"
field contained in [the package's "dc/schema.json"
file](https://github.com/ropenscilabs/deposits/blob/main/inst/extdata/dc/schema.json).
This schema file includes detailed specifications for every metadata term used
in `deposits`. The list of terms can also be viewed directly with [the
`dcmi_terms()`
function](https://docs.ropensci.org/deposits/reference/dcmi_terms.html). This
validation procedure is intended to issue instructive errors throughout when
metadata do not conform to the expected format, with this vignette
demonstrating several such error messages.

Let's start with a minimal version of the metadata list shown above, and try to
validate that using [the `deposit_fill_metadata()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-fill-metadata-):

```{r meta-validation1, eval = FALSE}
metadata <- list (
    author = "P.S. Reynolds",
    title = "Time-series analyses of beaver body temperatures."
)
cli <- depositsClient$new (service = "zenodo", sandbox = TRUE)
cli$deposit_fill_metadata (metadata)
```
```{r meta-validation1-error, echo = FALSE, eval = TRUE}
res <- data.frame (
    instancePath = "",
    schemaPath = "#/additionalProperties",
    keyword = "additionalProperties",
    additionalProperty = "author",
    message = "must NOT have additional properties"
)
print (res)
stop (
    "Stopping because the DCMI metadata terms listed ",
    "above do not conform with the expected schema.",
    call. = FALSE
)
```

That error tells us that "author" is an additional property, and that the
metadata "must NOT have additional properties."  These "additional properties"
refer to any beyond those specified in the schema. The schema used to validate
`deposits` metadata is contained in [the package's `dc/schema.json`
file](https://github.com/ropenscilabs/deposits/blob/main/inst/extdata/dc/schema.json).
A copy of that file is included with each installation of this package in the
location given by,

```{r schema-location}
system.file (fs::path ("extdata", "dc", "schema.json"), package = "deposits")
```

Searching for "author" either on the GitHub version or a local file will
quickly reveal that author information belongs in a field called "creator".
Renaming that metadata item then gives,

```{r meta-validation2, eval = FALSE}
metadata <- list (
    creator = "P.S. Reynolds",
    title = "Time-series analyses of beaver body temperatures."
)
cli$deposit_fill_metadata (metadata)
```
```{r meta-validation2-error, echo = FALSE, eval = TRUE}
res <- data.frame (
    instancePath = rep ("/creator", 3L),
    schemaPath = c (
        "#/properties/creator/anyOf/0/type",
        "#/properties/creator/anyOf/1/type",
        "#/properties/creator/anyOf"
    ),
    keyword = c ("type", "type", "anyOf"),
    type = c ("array", "array", NA_character_),
    message = c (
        rep ("must be array", 2L),
        "must match a schema in anyOf"
    ),
    required = c ("name", "name", NA_character_)
)
print (res)
stop (
    "Stopping because the DCMI metadata terms listed ",
    "above do not conform with the expected schema.",
    call. = FALSE
)
```

The first messages say that the "creator" property "must be array". Arrays in
JSON are `list` objects in R, so this suggests the next required modification:

```{r meta-validation3, eval = FALSE}
metadata <- list (
    creator = list ("P.S. Reynolds"),
    title = "Time-series analyses of beaver body temperatures."
)
cli$deposit_fill_metadata (metadata)
```
```{r meta-validation3-error, echo = FALSE, eval = TRUE}
res <- data.frame (
    instancePath = c (rep ("/creator/0", 2L), "/creator"),
    schemaPath = c (
        "#/properties/creator/anyOf/0/items/type",
        "#/properties/creator/anyOf/1/items/type",
        "#/properties/creator/anyOf"
    ),
    keyword = c ("type", "type", "anyOf"),
    type = c ("object", "object", NA_character_),
    message = c (
        rep ("must be object", 2L),
        "must match a schema in anyOf"
    ),
    required = c ("name", "name", NA_character_)
)
print (res)
stop (
    "Stopping because the DCMI metadata terms listed ",
    "above do not conform with the expected schema.",
    call. = FALSE
)
```

The items within the "creator" field must now be of type "object." R being the
relatively simply language that it is, "objects" are also `list`s, leading us
to:

```{r meta-validation4, eval = FALSE}
metadata <- list (
    creator = list (list ("P.S. Reynolds")),
    title = "Time-series analyses of beaver body temperatures."
)
cli$deposit_fill_metadata (metadata)
```
```{r meta-validation4-error, echo = FALSE, eval = TRUE}
res <- data.frame (
    instancePath = c (rep ("/creator/0", 2L), "/creator"),
    schemaPath = c (
        "#/properties/creator/anyOf/0/items/type",
        "#/properties/creator/anyOf/1/items/type",
        "#/properties/creator/anyOf"
    ),
    keyword = c ("type", "type", "anyOf"),
    type = c ("object", "object", NA_character_),
    message = c (
        rep ("must be object", 2L),
        "must match a schema in anyOf"
    ),
    required = c ("name", "name", NA_character_)
)
print (res)
stop (
    "Stopping because the DCMI metadata terms listed ",
    "above do not conform with the expected schema.",
    call. = FALSE
)
```

And that generates the same error. Looking a bit further in the error output
reveals that the object has a "required" property of "name". More generally,
for any instances in which the error messages themselves are not helpful, it is
useful to refer to the [actual JSON
schema](https://github.com/ropenscilabs/deposits/blob/main/inst/extdata/dc/schema.json)
(or a local version). The "creator" field expected for the Zenodo service can
then be seen to accept the properties, "name", "affiliation", "orcid", and
"gnd", with the "name" property being required, as indicated by the error
message.

```{r meta-validation5, eval = FALSE}
metadata <- list (
    creator = list (list (name = "P.S. Reynolds")),
    title = "Time-series analyses of beaver body temperatures."
)
cli$deposit_fill_metadata (metadata)
```
```{r meta-validation5-error, echo = FALSE, eval = TRUE}
res <- data.frame (
    instancePath = "/metadata",
    schemaPath = "#/properties/metadata/required",
    keyword = "required",
    missingProperty = "description",
    message = "must have required property 'description'"
)
print (res)
stop (
    "Stopping because the metadata terms listed above ",
    "do not conform with the expected schema for the ",
    "zenodo service.",
    call. = FALSE
)
```

That error seems to indicate that our "creator" field is then valid, yet
triggers a further error that that metadata "must have required property
'description'". Let's add that:

```{r meta-validation6, eval = FALSE}
metadata <- list (
    creator = list (list (name = "P.S. Reynolds")),
    title = "Time-series analyses of beaver body temperatures.",
    description = "Original source of 'beaver' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry."
)
cli$deposit_fill_metadata (metadata)
```

That now works, although that does not capture the full metadata shown above
from `?beaver1`. We still need information on the publisher. Examining the JSON
schema shows that it includes a "publisher" field which is expected to be a
string.

```{r meta-validation7, eval = FALSE}
metadata <- list (
    creator = list (list (name = "P.S. Reynolds")),
    title = "Time-series analyses of beaver body temperatures.",
    description = "Original source of 'beaver' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry.",
    publisher = "John Wiley and Sons"
)
cli$deposit_fill_metadata (metadata)
```

We may then wish to specify a date on which the resource was created. Searching
for "date" in the schema itself reveals that there is a "date" field, but it is
intended to describe, "A point or period of time associated with an event in
the lifecycle of the resource." There are various other "date" fields, but the
one we want is pretty clearly the "created" field, which is expected to be a
string in either "date" or "date-time" format. Let's try a simple "date":


```{r meta-validation8, eval = FALSE}
metadata <- list (
    creator = list (list (name = "P.S. Reynolds")),
    created = "1994-01-01",
    title = "Time-series analyses of beaver body temperatures.",
    description = "Original source of 'beaver' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry.",
    publisher = "John Wiley and Sons"
)
cli$deposit_fill_metadata (metadata)
```
```{r meta-validation8-error, echo = FALSE, eval = TRUE}
res <- data.frame (
    instancePath = "/created",
    schemaPath = "#/properties/created/format",
    keyword = "format",
    format = "date-time",
    message = 'must match format "date-time"'
)
print (res)
stop (
    "Stopping because the metadata terms listed above ",
    "do not conform with the expected schema for the ",
    "zenodo service.",
    call. = FALSE
)
```

That then fails because "created" must match the format "date-time", even
though the main JSON schema clearly indicates that either date or date-time
strings are acceptable. This error reflects the fact that validation of
`deposits` metadata is a two-stage process. The metadata are first validated
against the general JSON schema described above, then converted to formats
expected for the nominated deposits service, and the result is then validated
again against a service-specific JSON schema. In this case, the Zenodo service
itself requires "created" fields to have "date-time", and not "date", format.


```{r meta-validation9, eval = FALSE}
metadata <- list (
    creator = list (list (name = "P.S. Reynolds")),
    created = "1994-01-01T00:00:00",
    title = "Time-series analyses of beaver body temperatures.",
    description = "Original source of 'beaver' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry.",
    publisher = "John Wiley and Sons"
)
cli$deposit_fill_metadata (metadata)
```

That metadata is then finally in a valid format for the Zenodo service. An
instructive exercise is to change the service to "figshare", and use the
resultant error messages to tweak the metadata to the expected format for a
Figshare deposit.


### Metadata to specify connections between deposits

The schema permits many other fields, importantly including fields intended to
be used by the `deposits` package to document and identify connections between
different data resources. These include "hasPart", "hasVersion", "isPartOf",
"isReferencedBy", "isReplacedBy", "isRequiredBy", and "isVersionOf". Specifying
these allows one deposit to be connected to others. One of the aims of the
`deposits` package is to facilitate specification of these kinds of metadata,
and so to enhance an ability to inter-connect and inter-relate different data
depositions, providing insight into a growing ecosystem of inter-connected
deposits.

These "is" and "has" metadata fields have complex structures, but again the
JSON schemas can and should be used to understand the expected format. The
`isPartOf` field for Zenodo is expected to by an array (in R terms, a `list`)
of objects, each of which must have an "identifier" and "relation", and may
also have a "resource_type".

```{r metadata-isPartOf1, eval = FALSE}
metadata <- list (
    creator = list (list (name = "P.S. Reynolds")),
    created = "1994-01-01T00:00:00",
    title = "Time-series analyses of beaver body temperatures.",
    description = "Original source of 'beaver' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry.",
    publisher = "John Wiley and Sons",
    isPartOf = list (list (
        identifier = "Case Studies in Biometry",
        relation = "isPartOf"
    ))
)
cli$deposit_fill_metadata (metadata)
```

While those metadata are successfully validated, attempting to create that
deposit on Zenodo would trigger an error, because the "identifier" must conform
to an expected type. The JSON schema should as always be considered the
definitive reference, and that states for "identifier" that:

> "Supported identifiers include: DOI, Handle, ARK, PURL, ISSN, ISBN, PubMed ID, PubMed Central ID, ADS Bibliographic Code, arXiv, Life Science Identifiers (LSID), EAN-13, ISTC, URNs and URLs."

The "Case Studies in Biometry" book has an "ARK" identifier which can be used
here, 

```{r metadata-isPartOf2, eval = FALSE}
metadata <- list (
    creator = list (list (name = "P.S. Reynolds")),
    created = "1994-01-01T00:00:00",
    title = "Time-series analyses of beaver body temperatures.",
    description = "Original source of 'beaver' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry.",
    publisher = "John Wiley and Sons",
    isPartOf = list (list (
        identifier = "ark:/13960/t0mt2n370",
        relation = "isPartOf"
    ))
)
cli$deposit_fill_metadata (metadata)
```

Examining the JSON schema also reveals that "isPartOf" allows one additional
field of "resource_type", which has to be an upload, publication, or image type
followed by a type specification. These type specifications are included in the
second JSON schema used to validate Zenodo-specific metadata. The [GitHub
version can be seen
here](https://github.com/ropenscilabs/deposits/blob/main/inst/extdata/zenodo/schema.json),
or the location of the locally-installed version found with,

```{r sysfile2}
system.file (fs::path ("extdata", "zenodo", "schema.json"), package = "deposits")
```

This schema then reveals the accepted types for "publication" include "book",
so the "resource_type" becomes "publication-book":

```{r metadata-isPartOf3, eval = FALSE}
metadata <- list (
    creator = list (list (name = "P.S. Reynolds")),
    created = "1994-01-01T00:00:00",
    title = "Time-series analyses of beaver body temperatures.",
    description = "Original source of 'beaver' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry.",
    publisher = "John Wiley and Sons",
    isPartOf = list (list (
        identifier = "ark:/13960/t0mt2n370",
        relation = "isPartOf",
        resource_type = "publication-book"
    ))
)
cli$deposit_fill_metadata (metadata)
```

Finally we can confirm that those metadata are in a format which will be
accepted on our specific Zenodo service by creating a new deposit:

```{r create-deposit, eval = FALSE}
cli$deposit_new ()
#> ID of new deposit : 1186243
print (cli)
#> <deposits client>
#>  deposits service : zenodo
#>            sandbox: TRUE
#>          url_base : https://sandbox.zenodo.org/api/
#>  Current deposits : 1 (see 'deposits' element for details)
#>
#>  url_service : https://sandbox.zenodo.org/deposit/1186243
#>   deposit id : 1186243
#>     hostdata : list with 14  elements
#>     metadata : 6 terms (see 'metadata' element for details)
```

And that deposit has been successfully created, with the Zenodo record
populated with appropriately-translated versions of the metadata.

## Metadata templates

An alternative way to enter metadata into a `deposits` client is via a local
JSON metadata file. `depostis` includes the function,
[`deposits_metadata_template()`](https://docs.ropensci.org/deposits/reference/deposits_metadata_template.html),
which generates a local version of a complete template for all allowed metadata
fields. Note that deposits workflows are generally expected to use metadata
templates defined as part of [a "frictionless"
workflow](https://frictionlessdata.io), as described in detail in the
accompanying [frictionless
vignette](https://docs.ropensci.org/deposits/articles/frictionless.html). The 
[`deposits_metadata_template()`](https://docs.ropensci.org/deposits/reference/deposits_metadata_template.html)
is intended more for demonstration purposes, while a full frictionless template
is generally preferred. The format of deposits metadata nevertheless remains
the same in both.

```{r}
f <- fs::file_temp (ext = ".json")
deposits_metadata_template (filename = f)
```

That command issues an important note, given in even more detail in the first
line of the resultant template:

```{r}
head (readLines (f), 2L) [2]
```

A version of that template modified to reflect the metadata defined above would
then look like this:

```{r meta-template, echo = FALSE}
j <- '{
  "created": "1994-01-01T00:00:00",
  "creator": [
    {
      "name": "P.S. Reynolds"
    }
  ],
  "description": "Original source of \'beaver\' data, in Chapter 11 of  Lange, N., Ryan, L., Billard, L., Brillinger, D., Conquest, L. and Greenhouse, J. eds (1994) Case Studies in Biometry.",
  "isPartOf": [
      {
        "identifier": "ark:/13960/t0mt2n370",
        "relation": "isPartOf",
        "resource_type": "publication-book"
      }
  ],
  "publisher": "John Wiley and Sons",
  "title": "Time-series analyses of beaver body temperatures."
}'
writeLines (j, f)
j <- jsonlite::read_json (f)
jsonlite::toJSON (j, pretty = TRUE, auto_unbox = TRUE)
```

The format of each item within this JSON object then then matches the
specifications of the [main `dc/schema.json`
file](https://github.com/ropenscilabs/deposits/blob/main/inst/extdata/dc/schema.json).
Note that that schema expects both the "creator" and "isPartOf" fields for
Zenodo to be of type "array". Arrays in JSON are defined by square brackets, `[
... array contents ...]`. Both of these arrays are then expected to have items
of type "object", which in JSON are defined by curly brackets, `{ ... object
contents ...}`.

Those are the only two key aspects of JSON specification required to understand
and use metadata templates in `deposits`:

- A JSON "array" is defined within square brackets, and may contain any number
  of items: `[ ... array ...]`.
- A JSON "object" is defined within curly brackets, and must generally follow a
  defined structure: `{ ... object ...}`.

Both "array" and "object" items in JSON are represented in R terms as `list`s,
so an array of objects simply becomes `list(list(...))`, as demonstrated in the
first second of this vignette. (While this means that objects composed of
arrays could cause confusion, these are relatively uncommon, and are not used
in any of the JSON schemas within the `deposits`
package.)

### Valid JSON Files

The schema for the "creator" field shown above requires (for Zenodo) an "array"
of items, each of which is an "object" which must include a "name" field, and
may include additional fields for "affiliation", "orcid", and "gnd". An example
JSON specification for "creator" would thus be:

```{}
{
  "creator": [
    {
      "name": "A. Person",
      "orcid": "0000-0001-0002-0003"
    },
    {
      "name": "B. Person"
    }
  ]
}
```

Because "creator" is an "array", it may contain multiple unnamed items. In
contrast, items within an "object" must generally be named, and schemas will
generally expect single instances only of each item within an "object."
Switching the square and curly brackets in that definition would invalidate the
JSON file, and would trigger an error in `deposits`. 

Errors in the validation of JSON files themselves are, however, nowhere near as
informative as errors arising through comparing JSON files with schemas. If a
JSON file is itself invalid, the errors shown by `deposits` will generally be
uninformative. It is therefore very important to ensure that JSON metadata
files are formatted  in a valid way. The easiest way to check validity before
passing a metadata file to a `deposits` client is to use [the `validate()`
function of the `jsonlite`
package](https://cran.r-project.org/package=jsonlite). This function requires a
string input, so should be called as
`jsonlite::validate(readLines("metadata.json"))`.

This example demonstrates one important aspect of hand-writing JSON files,
which is commas. Commas **must** be used to separate any sequential fields, and
**must not** be placed at the end of any individual or sequence of fields. The
first statement means that valid JSON requires commas like this:

```{}
{
  "field1": "value1",
  "field2": "value2"
}
```

That specification without the comma would not be valid. The second statement
means that placing a comma after the second `"value2"` statement would also
render that JSON invalid. Note that these commas are equally important in
separating JSON objects:

```{}
{
  "object1": {
    "field": "value"
  },
  "object2": {
    "field": "value"
  }
}
```

That specification requires the single comma separating the two objects.
Placing commas anywhere else, such as after the values, or at the end of
"object2", would then invalidate the JSON.

## Metadata and the deposits frictionless workflow

As described in detail in [the "frictionless"
vignette](https://docs.ropensci.org/deposits/articles/frictionless.html),
`deposits` is designed to work seamlessly with [the "frictionless"
workflow](https://frictionlessdata.io/), via rOpenSci's ["frictionless" R
package](https://docs.ropensci.org/frictionless). For metadata, this means that
as soon as a data resource is uploaded to a service (with [the
`deposit_upload_file()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-upload-file-)),
a "datapackage.json" file is automatically created and filled with both
descriptions of the uploaded file, or "resource" in frictionless terms, as well
as the metadata described in this vignette. From that point onward, metadata
can be edited directly in the "datapackage.json" file, and both the local
client and external deposit can always be updated any time those local data
are modified with [the `deposit_update()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-update-).

In conclusion: from the moment your deposit has its own "datapackage.json"
file, that should be considered the definitive record of all metadata. Any
changes you may wish to make to those metadata should be made by directly
editing that file, and `deposits` will update all corresponding local and
remote records.

After restarting an R session, a deposits client can be connected to  a local
deposit by passing the path to the local directory containing deposit data and
a "datapackage.json" file as the "metadata" parameter of the `new()` method:

```{r frictionless-start1, eval = FALSE}
cli <- depositsClient$new (
    service = "<service>",
    metadata = "/<local>/<directory>/<or>/<frictionless_file>"
)
```

Equivalently, start an empty `deposits` client, and fill the metadata with [the
`deposit_fill_metadata()`
method](https://docs.ropensci.org/deposits/reference/depositsClient.html#method-deposit-fill-metadata-):

```{r frictionless-start2, eval = FALSE}
cli <- depositsClient$new (service = "<service>")
path <- "/<local>/<directory>/<or>/<frictionless_file>"
cli$deposit_fill_metadata (metadata = path)
```

The resultant client will fill the "metadata" field with the contents of the
"datapackage.json" file. If the deposit also exists remotely, any local changes
to "datapackage.json" will be reflected in changes on the remote deposits
service, including changes within the copy of "datapackage.json" held there.
